### 2022-02-15 writeup/paper.Rmd
first draft of writeup

---
title: "Estimating epidemiological quantities from repeated cross-sectional prevalence measurements"
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: false
  bookdown::html_document2: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE,
                      warning = FALSE, message = FALSE,
                      root.dir = here::here())

library(data.table)
library(purrr)
library(here)
library(kableExtra)
## Get tools
functions <- list.files(here("R"), full.names = TRUE)
walk(functions, source)

params_table <- here::here("outputs", "example-parameters.csv")
if (!file.exists(params_table)) {
  source(here::here("scripts", "simple-example.R"))
}
params <- fread(params_table)
params[name == "gamma", name := ifelse(p_index == 1, "gamma (infection)", "gamma (vaccination)")]
params[, Description := fcase(
           name == "beta", "Proportion infected that seroconvert",
           name == "gamma (infection)", "Antibody waning following infection",
           name == "gamma (vaccination)", "Antibody waning following vaccination",
           name == "delta", "Proportion vaccinated that seroconvert",
           name == "k", "Efficacy adjustment of immunity following infection",
           name == "l", "Efficacy adjustment of immunity following vaccination")]
biol_params <- params[!is.na(Description)]
biol_params[, `Estimate (90% CI)` := paste0(signif(q5, 2), "--", signif(q95, 2))]
biol_params <- biol_params[, list(Parameter = name,
                                  Description,
                                  `Estimate (90% CI)`)]

```

# Introduction

Infectious disease surveillance serves to monitor the health of populations and identify new threats as quickly as possible after they arise [@murray2017].
It is often based on healthcare-based reporting systems whereby primary care providers or hospitals report numbers of individuals identified as likely cases of a disease to central authorities where these numbers are collated and reported as aggregates.
During the Covid-19 pandemic in the United Kingdom, reporting of cases has mostly involved collating numbers of laboratory-identified infections with SARS-CoV-2 via self-reporting, community testing sites or hospitals.

A separate and independent system of collating information on the state of the pandemic has been run by the Office for National Statistics (ONS) via its Community Infection Survey, which conducts repeated cross-sectional surveys of Polymerase Chain Reaction (PCR) positivity indicating infection with SARS-CoV-2, as well as antibody seroprevalence via household visits [@pouwels2020]. 
By adjusting for biases in the sampled population, the study has been used to estimate daily population-wide estimates of infection prevalence, unaffected by testing capacity or reporting behaviour that often varies by age as well as sociodemographic or other factors.

While repeated randomised cross-sectional sampling of positivity and antibodies provides utility in themselves for tracking an epidemic in real time, they can also be used for estimating epidemiological quantities by combining them with information on infection kinetics and immunological responses.
Here we present a semi-mechanistic model that combines PCR positivity curves, generation interval estimates and vaccination data with ONS PCR positivity and antibody data to estimate infection incidence and its growth rate, reproduction numbers and rates of antibody waning.

# Methods

## Data

We obtained the published estimates of daily prevalence of Polymerase Chain Reaction (PCR) positivity beginning on 26 April, 2020, from the ONS Community infection survey separately by nation, region, age group and variant, alongside their 95% confidence intervals, from the published spreadsheets on the ONS web site.
ONS estimates of a given prevalence vary between publication dates as the internal model to calculate prevalence involves smoothing, such that new data points in the present affect the estimates of times past.
We aggregated estimates of PCR positivity for a single day produced for different publication dates by calculating the central estimate and confidence limits as the medians of the different respective central estimates and confidence limits.

## Model

We developed Bayesian model to estimate epidemiological quantities from ONS PCR positivity estimates and, optionally, population level antibody prevalence estimates and vaccination coverage.

### PCR positivity

We estimated the population proportion newly infected in the population $I(t)$ as a latent variable that is convolved with an PCR positivity curve $p(s)$, the probability of someone infected at time $s=0$ to test PCR positive to yield prevalence of PCR positivity $P(t)$.
$$
P(t) = \sum_{s= 0}^{t_\text{p,max}} p(s) I(t - s)
$$
where $t_\text{p,max}=60$ is the maximum time modelled for which a person can stay PCR positive.
We assumed each $p(s)$ to have an independent normal prior distribution at each time $s$ after infection with given mean and standard deviation estimates from another study [@hellewell2021]. 
Infection incidence $I(t)$ is distinct form the estimates of PCR positivity incidence provided by ONS alongside the prevalence estimates, as it allows for the probability of infections testing yielding negative PCR results as a function of the time since infection.

We used Gaussian Process (GP) priors to ensure smoothness of the estimates and deal with data gaps, whereby alternatively either $I(t)$ is has a GP prior with exponential quadratic kernel.
$$
\begin{aligned}
I(t) &\sim \text{logit}(i_0 + i(t))\\
i(t) &\sim \text{GP}(t)
\end{aligned}
$$
where $i_0$ is the estimated mean of the GP, or the GP prior is applied to higher order differences, for example the growth rate such as
$$
i(t) - i(t - 1) \sim \text{GP}(t)
$$
which implies that growth, rather than incidence, remains at an estimated mean level in the absence of data, usually leading to better real-time performance [@abbott2020].
The results shown in this paper were obtained using this formulation with a GP prior on the growth rate.

We assumed that the probability of observing prevalence $Y_{\text{P}, t}$ at time $t$ was given by independent normal distributions with mean $P(t)$ and standard deviation 
$$\sigma_{\text{P}, t} = \sigma_\text{P} + Y^\sigma_{\text{P}, t}$$
where $\sigma_\text{P}$ was estimated as part of the inference procedure and $Y^\sigma_{\text{P}, t}$ calculated based on the reported confidence intervals in the ONS data, assuming independent normal errors.
For data sets where only weekly estimates were reported by ONS, for example at the sub-regional level, we calculated average prevalence across the time period reported from our daily prevalence estimates.

Using the estimate infection incidences $I(t)$ we estimated growth rates $r(t)$ as
$$
r(t) = \log I(t) - \log I(t - 1)
$$
and reproduction numbers $R(t)$ using the renewal equation as
$$
R(t) = \frac{I(t)}{\sum_{s=0}^{t_\text{g,max}} g(s) I(t - s)}
$$
where $g(s)$ is the distribution of the generation interval since the time of infection [@fraser2007].
We assumed a maximum generation interval of $t_\text{g,max}=14$.
We use re-estimated generation intervals from early in the pandemic in Singapore as reported previously [@abbott2020].

### Antibodies

When additionally using antibodies we convolve the modelled infections $I(t)$ as well as input data on vaccinations $Y_{\text{V}, t}$ with distributions quantifying the delay to generating detectable antibodies following infection (by default set to 4 weeks for both infection and vaccination), yielding potentially antibody-generating time series from infection $I^{\text{A}}$ and $V^{\text{A}}$.
We then calculate antibodies from infection as
$$
A^{\text{I}}(t) =
A^{\text{I}}(t - 1) +
\beta I^{\text{A}}(t)  (1 - A(t - 1))^k -
\gamma_\text{I} A^{\text{I}}(t - 1) 
$$
and antibodies from vaccination as
$$
A^{\text{V}}(t) =
A^{\text{V}}(t - 1) +
\delta V^{\text{A}}(t) (1 - A(t - 1))^l -
 \gamma_\text{V} A^{\text{V}}(t - 1)
$$
with the total population proportion with antibodies given as the sum of the two,
$$
A(t) = A^{\text{I}}(t)  + A^{\text{V}}(t)  
$$

Here, the additional parameter $\beta$ can be interpreted as proportion of new infections that does not increase the population proportion with antibodies, either due to lack of seroconversion or because they are breakthrough infections in those with existing antibodies, and parameters $k$ and $l$ govern the degree to which new seropostives preferentially arise in those not seropositive so far. 
Additional parameters $\gamma_\text{I}$ and $\gamma_\text{V}$ can be interpreted as rates of waning from natural infection and vaccination, respectively.
This formulation implies simplifying assumptions that the rate of waning of detectable antibodies is exponential, that vaccine doses are allocated randomly amongst those with or without existing antibodies, and that the proportion of new vaccinations that lead to seroconversion $\delta$ is constant and independent of age, vaccine use, and dose number.

## Implementation

The model was implemented in _stan_ and using the _cmdstanr_ R package. 
All code needed to reproduce the results shown in this paper is available at https://github.com/epiforecasts/inc2prev.

# Results

```{r prevalence-plot, out.width=".7\\paperwidth", fig.cap = "Model posteriors for England. A. Estimates of daily modelled prevalence and modelled prevalence as published by ONS. B. Estimated incidence of new infections. C. Estimated antibody prevalence and estimes as published by ONS. D. Estimated reproduction numbers."}
knitr::include_graphics(here::here("figures", "readme", "example-estimates.png"))
```

The model was able to reproduce the daily prevalence estimates and weekly antibody prevalence estimates published by ONS with reasonable accuracy when run until 15 November 2021 (Figure \@ref(fig:prevalence-plot)).
The peaks of the corresponding incidence curve are earlier, higher and sharper.
Estimated reproduction numbers highlight some key phases of the UK pandemic between April 2020 and November 2021, in particular rapid increases due to emergence of the Alpha variant in December followed by a period of low transmission during lockdown until March 2021, and rapid spread of the Delta variant in May-July 2021 followed by a period of relatively steady transmission.

```{r params-table}
biol_params %>%
  kbl(caption = "90% credible intervals (as quantiles of the posterior distribution) of biological parameters.",
      booktabs = TRUE)
```

Posterior estimates of recovered biological parameters are shown in Table \@ref{tab:prams-table}.
Some of the parameter estimates show high levels of correlation suggesting issues of identifiability (Figure \@ref(fig:pairs-plot)).

```{r pairs-plot, out.width=".7\\paperwidth", fig.cap = "Scatter plots and histograms of posterior parameter samples."}
knitr::include_graphics(here::here("figures", "readme", "pairs.png"))
```

# Discussion

We have presented a method to estimate epidemiological parameters such as infection incidence, time-varying reproduction numbers and growth rates from repeated cross-sectional PCR positivity estimates reported.
The estimates of infection incidence are distinct from estimates of PCR positivity incidence that are reported alongside the positivity prevalence estimates, as the probability of detecting infections is low early in the course of an infection, and more generally varies over said course [@hellewell2021].
When additionally using antibody and vaccination data, we recover estimates of relevant parameters such as seroconversion and waning rates that can be used to estimate antibody prevalence where infection and vaccination data is available but antibody data is not.

As currently implemented, our method suffers from a number of limitations that risk biasing the results.
Several of the key parameters in our model, especially the estimates of PCR positivity over time from infection and generation interval distributions, are fixed and based on estimates derived from wildtype virus in a particular cohort of healthcare workers and may well be incorrect for other circulating variants or populations.
Furthermore, generation times have been shown to change over time due to behavioural changes and epidemiological dynamics, which would affect our reproduction number estimates [@champredon2015; @hart2021; @park2021].
PCR detection probabilities as a function of time since infection were based on independent normal distributions, whereas in reality they are likely to be correlated over time.
We modelled the growth of infections as a stationary Gaussian process, whereas in reality variation over time has changed between periods of stability and rapid change due to changes in contact behaviour in response to the epidemic.
Lastly, we assumed that antibody waning was exponential and conferred either full protection if with detectable antibodies, or none whatsoever if not, and ignored any consequences of multiple rounds of vaccination or infection apart from converting those without detectable antibodies to having detectable antibodies.

Future directions of this work should help address some of these limitations, for example by including more detail on antibody levels, or by including antibody measurements that may be able to distinguish between natural and vaccine-acquired immunity [@amjadi2021].
It could further make use of more comprehensive information on PCR detection curves taking into account correlations in detectability since time from infection.
Combined with other data streams, for example on test-positive community cases, or severe outcomes resulting in hospitalisations or deaths, our method could be used to understand rates of notification or sever disease given infection, or to generate forecasts of expected burden.
Lastly, more detailed information on the infections detected, for example viral loads via Cycle threshold (Ct) values, could be used to improve real-time performance of growth rates and reproduction numbers [@hay2021].

There is enormous potential for understanding epidemiological dynamics from repeated cross-sectional surveys.
Where the generation interval distribution is the same or close to the distribution of detectability after infection, this could be done using recently developed methods for unified modelling of incidence and prevalence [@pakkanen2021].
The methods presented here and related ones could be applied to other infections monitored in a similar way, and thus in combination with such data collection and publication become a tool for monitoring epidemic and endemic infectious diseases in the future.

# Acknowledgements

We thank Thomas House for insightful comments on this work, and the Office for National Statistics for making the data sets publicly available.

# References

### 2022-02-15 writeup/paper.Rmd
add authors

@@
 title: "Estimating epidemiological quantities from repeated cross-sectional prevalence measurements"
+author: Sam Abbott, Sebastian Funk
 bibliography: references.bib
 csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
-date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
+date: "`r format(Sys.Date(), format = '%B %d, %Y')` - WORK IN PROGRESS DRAFT"
 output:

### 2022-02-16 R/model.R
update prior

@@ -13,7 +13,7 @@ i2p_data <- function(prev, ab, vacc, init_ab,
                      init_cum_infections = c(0, 0),
                      inf_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
                      vacc_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
-                     prop_dont_seroconvert = c(-2, 1), # 10%
+                     prop_seroconvert = c(2, 1), # 90%
                      inf_waning_rate = c(-9, 4),
                      vac_waning_rate = c(-9, 4), # 0.1%
                      vaccine_efficacy = c(3, 1), # 95%
@@ -124,7 +124,7 @@ i2p_data <- function(prev, ab, vacc, init_ab,
     prob_detect_sd = rev(prob_detectable$sd),
     pbt = max(prob_detectable$time) + 1,
     init_inc_mean = logit(init_inc_mean),
-    pbeta = prop_dont_seroconvert,
+    pbeta = prop_seroconvert,
     pgamma_mean = c(inf_waning_rate[1], vac_waning_rate[1]),
     pgamma_sd = c(inf_waning_rate[2], vac_waning_rate[2]),
     pdelta = vaccine_efficacy,

### 2022-02-16 scripts/simple-example.R
example fixes

@@ -59,8 +59,8 @@ fit <- incidence(
   joint_data$vaccination[[1]],
   joint_data$initial_antibodies[[1]],
   variables = c(
-    "est_prev", "infections", "dab", "dcases", "r", "R",
-    "alpha", "beta", "gamma", "delta", "k", "l"
+    "est_prev", "infections", "dcases", "r", "R",
+    "est_ab", "dab", "beta", "gamma", "delta", "k", "l"
   ),
   prob_detect = prob_detect, parallel_chains = 2, iter_warmup = 200,
   chains = 2, model = mod, adapt_delta = 0.85, max_treedepth = 15,
@@ -97,7 +97,7 @@ stanfit <- read_stan_csv(fit$fit[[1]]$output_files())
 np <- nuts_params(stanfit)
 pairs <- mcmc_pairs(fit$fit[[1]]$draws(),
   np = np,
-  pars = c("alpha[1]", "beta[1]", "gamma[1]", "gamma[2]", "delta[1]", "k[1]", "l[1]")
+  pars = c("beta[1]", "gamma[1]", "gamma[2]", "delta[1]", "k[1]", "l[1]")
 )
 ggsave("figures/readme/pairs.png", pairs, width = 16, height = 16)
 

### 2022-02-17 writeup/paper.Rmd
fix figure

@@
 params <- fread(params_table)
-params[name == "gamma", name := ifelse(p_index == 1, "gamma (infection)", "gamma (vaccination)")]
+params[name == "gamma", name := ifelse(index == 1, "gamma (infection)", "gamma (vaccination)")]
 params[, Description := fcase(

### 2022-02-17 scripts/combine_results.R
run and combine

@@ -1,0 +1,24 @@
+library(data.table)
+library(tidyverse)
+
+child_age_groups = c('2-10', '11-15')
+estimates_c = readRDS('outputs/estimates_kids_combined_age_ab.rds')
+estimates_a = readRDS('outputs/estimates_age_ab.rds')
+
+estimates_c = estimates_c %>% filter(variable %in% child_age_groups)
+estimates_a = estimates_a %>% filter(!(variable %in% child_age_groups))
+
+estiamtes_all = rbind(estimates_c, estimates_a)
+
+saveRDS(estiamtes_all, 'outputs/estimates_combined_age_ab_A.rds')
+
+
+samples_c = readRDS('outputs/samples_kids_combined_age_ab.rds')
+samples_a = readRDS('outputs/samples_age_ab.rds')
+
+samples_c = samples_c %>% filter(variable %in% child_age_groups)
+samples_a = samples_a %>% filter(!(variable %in% child_age_groups))
+
+samples_all = rbind(samples_c, samples_a)
+
+saveRDS(samples_all, 'outputs/samples_combined_age_ab_A.rds')

### 2022-02-17 scripts/estimate.R

@@ -247,7 +247,7 @@ if (kids_ab) {
   pre_fit_ab_dat = list(n_ab_draws = dim(ab_draws)[1], n_ab_pars = dim(ab_draws)[2], ab_draws = ab_draws)
 
 
-  mod <- i2p_model("stan/inc2prev_prefitab_data.stan")
+  mod <- i2p_model("stan/inc_2_prev_abnodata.stan")
   # Run model fits in parallel
   plan(callr, workers = future::availableCores())
   est_kids <- future_lapply(

### 2022-02-21 R/incidence.R
Fit multiple datasets simultaneously

@@ -6,8 +6,7 @@
 #' @param ab Observed antibody prevalence data
 #' @param vacc Observed vaccination data
 #' @param init_ab Observed initial antibody data
-#' @param prob_detection A dataframe of posterior samples for the probability
-#' of detection
+#' @param prob_detect Time-varying probability of detection
 #' @param data_args A list of arguments to pass to `i2p_data()`
 #' @param model A stan model object as produced by `i2p_model()`
 #' or a similar compiled stan model.
@@ -15,6 +14,9 @@
 #' @param quantiles A vector of quantiles to return in summarising
 #' @param samples The number of samples to return
 #' @param keep_fit Whether to retain the stan fit object
+#' @param var_col Column of a common name in the data frames passed as data
+#' that should be treated as indicator of multiple data sets that should be used
+#' jointly for estimation
 #' @param p A `progressr` function used when fitting multiple models
 #' to track progress
 #' @param ... Additional arguments passed to `fit_fn`.
@@ -27,6 +29,7 @@ incidence <- function(prev, ab = NULL, vacc = NULL, init_ab = NULL, prob_detect,
                       variables = c(
                         "est_prev", "infections", "dcases", "r", "R"
                       ),
+                      var_col = NULL,
                       quantiles = seq(0.05, 0.95, by = 0.05),
                       samples = 100,
                       keep_fit = FALSE,
@@ -38,7 +41,8 @@ incidence <- function(prev, ab = NULL, vacc = NULL, init_ab = NULL, prob_detect,
         ab = ab,
         vacc = vacc,
         init_ab = init_ab,
-        prob_detectable = prob_detect
+        prob_detectable = prob_detect,
+        var_col = var_col
       ),
       data_args
     )
@@ -54,11 +58,13 @@ incidence <- function(prev, ab = NULL, vacc = NULL, init_ab = NULL, prob_detect,
 
   fit[, summary := list(
     i2p_summarise(fit[[1]], variables = variables, quantiles = quantiles) |>
+      i2p_add_var(prev = prev, data = dat, var_col = var_col) |>
       i2p_add_date(prev = prev, ab = ab, data = dat)
   )]
 
   fit[, samples := list(
     i2p_draws(fit[[1]], variables = variables, samples = ..samples) |>
+      i2p_add_var(prev = prev, data = dat, var_col = var_col) |>
       i2p_add_date(prev = prev, ab = ab, data = dat)
   )]
   if (!keep_fit) {

### 2022-02-21 R/postprocess.R

@@ -7,7 +7,9 @@ i2p_draws <- function(fit, variables = NULL, samples = 100) {
     dplyr::select(-.chain, -.iteration, -.draw) %>%
     tidyr::pivot_longer(matches("[0-9]")) %>%
     dplyr::mutate(
-      index = as.integer(sub("^.*\\[([0-9]+)]$", "\\1", name)),
+      n_index = as.integer(sub("^.*\\[([0-9]+),.+]$", "\\1", name)),
+      t_index = as.integer(sub("^.*\\[.+,([0-9]+)]$", "\\1", name)),
+      p_index = as.integer(sub("^.*\\[([0-9]+)]$", "\\1", name)),
       name = sub("\\[.*$", "", name)
     )
 }
@@ -20,7 +22,9 @@ i2p_summarise <- function(fit, variables = NULL,
     dplyr::as_tibble() %>%
     dplyr::rename(name = variable) %>%
     dplyr::mutate(
-      index = as.integer(sub("^.*\\[([0-9]+)]$", "\\1", name)),
+      n_index = as.integer(sub("^.*\\[([0-9]+),.+]$", "\\1", name)),
+      t_index = as.integer(sub("^.*\\[.+,([0-9]+)]$", "\\1", name)),
+      p_index = as.integer(sub("^.*\\[([0-9]+)]$", "\\1", name)),
       name = sub("\\[.*$", "", name)
     )
 }
@@ -36,11 +40,30 @@ i2p_add_date <- function(dt, prev, ab, data) {
       ,
       date := fcase(
         name %in% c("infections", "dcases", "dab"),
-        index - 1 + start_date - ut,
-        name == "est_prev", prev$date[index],
-        name == "est_ab", ab$date[index],
-        name == "r", index - 1 + start_date,
-        name == "R", index - 1 + start_date
+        t_index - 1 + start_date - ut,
+        name == "est_prev", prev$date[t_index],
+        name == "est_ab", ab$date[t_index],
+        name == "r", t_index - 1 + start_date,
+        name == "R", t_index - 1 + start_date
+      )
+    ]
+  )
+  return(dt[])
+}
+
+## translate index into variable
+i2p_add_var <- function(dt, prev, data, var_col = NULL) {
+  if (is.null(var_col)) return(dt[])
+  vars <- unique(rownames(data$prev))
+  ab_index <- data$ab_index
+
+  dt <- suppressWarnings(
+    data.table::as.data.table(dt)[
+      ,
[…]

### 2022-02-21 scripts/estimate.R

@@ -107,9 +107,38 @@ if (antibodies) {
     filter_opt(start_date) %>%
     nest(initial_antibodies = c(-variable))
   data <- data %>%
-    inner_join(ab, by = "variable") %>%
-    inner_join(vacc, by = "variable") %>%
-    inner_join(early, by = "variable")
+    left_join(ab, by = "variable") %>%
+    left_join(vacc, by = "variable") %>%
+    left_join(early, by = "variable")
+}
+
+## determine estimation strategy
+if (antibodies) {
+  if (regional) {
+    ## fit all at once
+    data <- data %>%
+      mutate(grouping = "all")
+  } else if (local) {
+    ## estimate parameters from regions
+  } else if (age) {
+    ## fit age groups without data alongside older ones
+    data <- data %>%
+      mutate(grouping = variable,
+	     grouping = recode(grouping,
+			       `2-10` = "16-24",
+			       `11-15` = "16-24"))
+  } else if (variants) {
+    ## doesn't really make sense with antibodies
+    stop("Fitting variant antibodies does not really make sense")
+  } else {
+    ## national: fit separately as time series have different lengths
+    data <- data %>%
+      mutate(grouping = variable)
+  }
+} else {
+  ## fit separately
+  data <- data %>%
+    mutate(grouping = variable)
 }
 
 if (local && !weekly) {
@@ -118,7 +147,7 @@ if (local && !weekly) {
 }
 
 data <- data %>%
-  group_split(variable)
+  group_split(grouping, .keep = FALSE)
 
 # Location probability of detection posterior
 prob_detect <- read_prob_detectable()
@@ -154,15 +183,19 @@ incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, week
 
   safe_incidence <- purrr::safely(incidence)
 
-  prev <- data$prevalence[[1]]
+  prev <- data %>%
+    unnest(prevalence)
   if (weekly) {
     prev <- prev %>%
       convert_to_weekly(max(.$end_date))
   }
   if (antibodies) {
-    ab <- data$antibodies[[1]]
-    vacc <- data$vaccination[[1]]
-    init_ab <- data$initial_antibodies[[1]]
+    ab <- data %>%
+      unnest(antibodies)
+    vacc <- data %>%
+      unnest(vaccination)
+    init_ab <- data %>%
+      unnest(initial_antibodies)
   } else {
     ab <- NULL
     vacc <- NULL
[…]

### 2022-02-21 scripts/update.sh

@@ -4,9 +4,11 @@ git pull -Xours
 
 Rscript data-raw/update_cis.R
 
-echo Space
+echo National
 git status | grep -q cis.csv && Rscript scripts/estimate.R -d 1 && Rscript scripts/analyse.R
-git status | grep -q cis.csv && Rscript scripts/estimate.R -n -r -d 1 && Rscript scripts/analyse.R -r
+
+echo Regional
+git status | grep -q cis.csv && Rscript scripts/estimate.R -r -n -d 1 && Rscript scripts/analyse.R -r
 
 echo Age
 git status | grep -q cis_age.csv && Rscript scripts/estimate.R -g -d 1 && Rscript scripts/analyse.R -g

### 2022-02-21 R/incidence.R
add model changes from #27

@@ -26,10 +26,8 @@
 incidence <- function(prev, ab = NULL, vacc = NULL, init_ab = NULL, prob_detect,
                       data_args = list(),
                       model = i2p_model(),
-                      variables = c(
-                        "est_prev", "infections", "dcases", "r", "R"
-                      ),
                       var_col = NULL,
+                      variables = NULL,
                       quantiles = seq(0.05, 0.95, by = 0.05),
                       samples = 100,
                       keep_fit = FALSE,

### 2022-02-21 R/postprocess.R

@@ -63,7 +63,8 @@ i2p_add_var <- function(dt, prev, data, var_col = NULL) {
       paste(var_col) := fcase(
         name %in% c("est_prev", "est_ab", "infections",
 		    "dcases", "dab", "r", "R"), vars[n_index],
-        name %in% c("beta", "gamma", "delta"), paste0(vars[ab_index], collapse = ";")
+        name %in% c("beta", "gamma", "delta", "k", "l"),
+        paste0(vars[ab_index], collapse = ";")
       )
     ]
   )

### 2022-02-21 R/incidence.R
update incidence description

@@ -11,15 +11,14 @@
 #' @param data_args A list of arguments to pass to `i2p_data()`
 #' @param model A stan model object as produced by `i2p_model()`
 #' or a similar compiled stan model.
-#' @param inits_fn A function that returns a function which produces
-#' a set of initial conditions. Defaults to `i2p_inits()`.
-#' @param fit_fn A function used to fit the `model` object. Defaults
-#' to `i2p_sample`.
+#' @param variables A vector of variables to summarise and return
+#' @param quantiles A vector of quantiles to return in summarising
+#' @param samples The number of samples to return
+#' @param keep_fit Whether to retain the stan fit object
 #' @param p A `progressr` function used when fitting multiple models
 #' to track progress
 #' @param ... Additional arguments passed to `fit_fn`.
-#' @return A fit stan model as produced by the option passed to
-#' `fit_fn`.
+#' @return A data.table including summarised estimates and parameter samples.
 #' @export
 #' @family incidence
 incidence <- function(prev, ab = NULL, vacc = NULL, init_ab = NULL, prob_detect,

### 2022-02-21 scripts/analyse.R
fix typo in analyse script

@@ -43,7 +43,7 @@ walk(functions, source)
 
 if (local) {
   suffix <- "_local"
-if (regional) {
+} else if (regional) {
   suffix <- "_regional"
 } else if (age) {
   suffix <- "_age"

### 2022-02-21 R/plot_wrapper.R
define early ab

@@ -32,7 +32,8 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
     cumulative_infection_samples <- level_samples %>%
       filter(name == "cumulative_infections")
     n_samples <- max(cumulative_infection_samples$sample)
-    early_samples <- level_early %>%
+    early_samples <- early %>%
+      filter(level == {{ level }}) %>%
       group_by(variable) %>%
       summarise(
         rand = list(tibble(

### 2022-02-21 R/model.R
remove superfluous code

@@ -10,7 +10,6 @@ i2p_gp_tune_model <- function(path) {
 # define required stan data
 i2p_data <- function(prev, ab, vacc, init_ab,
                      prob_detectable, unobserved_time = 14, horizon = 0,
-                     init_cum_infections = c(0, 0),
                      inf_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
                      vacc_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
                      prop_dont_seroconvert = c(-2, 1), # 10%
@@ -165,7 +164,6 @@ i2p_data <- function(prev, ab, vacc, init_ab,
   if (is.na(gp_ls[2])) {
     gp_ls[2] <- dat$t
   }
-  gp_ls
   lsp <- tune_inv_gamma(gp_ls[1], gp_ls[2], gp_tune_model)
   dat$lengthscale_alpha <- lsp$alpha
   dat$lengthscale_beta <- lsp$beta

### 2022-02-21 scripts/estimate.R
add option to aggregate data to weekly

@@ -32,6 +32,7 @@ Options:
     -n, --nhse                 Analyse NHSE regions
     -d, --differencing=<level> Level of differencing of GP (0 = infections,  1 = growth, 2 = differences in growth etc.)
     -e, --start_date=<date>    Start date to use for estimation
+    -w, --weekly               Aggregate all data to weekly
     -c, --gp_frac=<frac>       Fraction of latent timepoints to use in the Gaussian
                                process approximation. Reducing this improves runtimes at
                                the cost of reducing the accuracy of the Gaussian process
@@ -53,6 +54,7 @@ variants <- !is.null(opts$variants) && opts$variants
 nhse <- !is.null(opts$nhse) && opts$nhse
 differencing <- ifelse(is.null(opts$differencing), 0L, as.integer(opts$differencing))
 start_date <- as.Date(opts$start_date)
+weekly <- !is.null(opts$weekly) && opts$weekly
 gp_frac <- ifelse(is.null(opts$gp_frac), 0.3, as.numeric(opts$gp_frac))
 
 ## Get tools
@@ -110,6 +112,11 @@ if (antibodies) {
     inner_join(early, by = "variable")
 }
 
+if (local && !weekly) {
+  warning("Converting everything to weekly as needed for local estimates")
+  weekly <- TRUE
+}
+
 data <- data %>%
   group_split(variable)
 
@@ -125,9 +132,24 @@ tune <- i2p_gp_tune_model()
 ## Fit model
 dir.create(here::here("outputs"), showWarnings = FALSE)
 
+convert_to_weekly <- function(x, ref_end_date, cols = c("middle", "lower", "upper"), aggregate = mean) {
+  if (!("end_date" %in% colnames(x))) {
+    x <- x %>%
+      mutate(start_date = date,
+	     end_date = date)
+  }
+  x %>%
+    pivot_longer(all_of(cols)) %>%
+    mutate(end_date = end_date + as.integer(ref_end_date - end_date) %% 7) %>%
+    group_by(across(c(-value, -start_date, -date))) %>%
+    summarise(start_date = min(start_date), value = aggregate(value), .groups = "drop") %>%
+    mutate(date = start_date + (end_date - start_date) / 2) %>%
+    pivot_wider()
+}
+
 # create a helper function to estimate the model and apply some
 # summary statistics
-incidence_with_var <- function(data, pb, model, gp_model, differencing = 0) {
[…]

### 2022-02-21 scripts/estimate.R
pass additional options to incidence

@@ -182,7 +182,10 @@ incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, week
     variables = variables,
     prob_detect = pb, parallel_chains = 2, iter_warmup = 250,
     chains = 2, model = mod, adapt_delta = 0.9, max_treedepth = 12,
-    data_args = list(gp_tune_model = gp_model, gp_m = gp_frac, differencing = differencing)
+    data_args = list(
+      gp_tune_model = gp_model, gp_m = gp_frac, differencing = differencing
+    ),
+    ...
   )
 
   if (is.null(fit$result)) {

### 2022-02-21 scripts/estimate.R
pass on differencing argument

@@ -211,6 +211,7 @@ est <- future_lapply(
   data, incidence_with_var,
   pb = prob_detect,
   model = mod, gp_model = tune,
+  differencing = differencing,
   future.seed = TRUE
 )
 

### 2022-02-21 scripts/plot_estimates.r
improve plots of estimates

@@ -7,15 +7,26 @@ library(readr)
 library(scales)
 library(tidyr)
 
+source(here::here("R", "read.R"))
+prev <- read_cis()
+local_region <- prev %>%
+  filter(level == "local") %>%
+  select(level, variable, region) %>%
+  distinct()
+
+dir.create(here::here("figures", "additional"),
+           showWarnings = FALSE, recursive = TRUE)
+
 ## Get tools
 files <-
-  list.files(here("outputs"), pattern = "estimates.*\\.csv", full.names = TRUE)
+  list.files(here("outputs"), pattern = "estimates_[^_]+\\.csv", full.names = TRUE)
 data <- map(files, read_csv) %>%
   bind_rows() %>%
   separate(variable, c("variable", "geography"), sep = "\\|")
+data <- left_join(data, local_region, by = c("level", "variable"))
 
 cum_files <-
-  list.files(here("outputs"), pattern = "cumulative.*\\.rds", full.names = TRUE)
+  list.files(here("outputs"), pattern = "cumulative_[^_]+\\.rds", full.names = TRUE)
 if (length(cum_files) > 0) {
   cum_data <- map(cum_files, readRDS) %>%
     bind_rows() %>%
@@ -43,14 +54,15 @@ if (length(cum_files) > 0) {
 }
 
 data <- data %>%
-  mutate(variable = if_else(variable == "2-10", "02-10", variable))
+  mutate(variable = if_else(variable == "2-10", "02-10", variable),
+         name = if_else(name == "R", "Rt", name))  
 
 var_names <- c(
   est_prev = "Prevalence estimate",
   infections = "Estimated incidence",
   dcases = "Daily estimated prevalence",
   r = "Growth rate",
-  R = "Reproduction number",
+  Rt = "Reproduction number",
   cumulative_infections = "Cumulative attack rate"
 )
 
@@ -59,7 +71,7 @@ labels <- list(
   infections = scales::comma,
   dcases = scales::percent_format(1L),
   r = waiver(),
-  R = waiver(),
+  Rt = waiver(),
   cumulative_infections = scales::percent_format(1L)
 )
 
@@ -79,12 +91,15 @@ breaks <- c(all = "2 months", `1year` = "1 month", `3months` = "1 month")
 
 for (history in names(histories)) {
   for (level in unique(non_variant$level)) {
+    colour_var <- ifelse(level == "local", "region", "variable")
     for (name in unique(non_variant$name)) {
       p <- ggplot(non_variant %>%
[…]

### 2022-02-21 scripts/update.sh

@@ -13,6 +13,11 @@ git status | grep -q cis_age.csv && Rscript scripts/estimate.R -g -d 1 && Rscrip
 echo Variants
 git status | grep -q cis_variants.csv && Rscript scripts/estimate.R -v -d 1 && Rscript scripts/analyse.R -v
 
+Rscript scripts/plot_estimates.r
+mv figures/additional/national_infections_all.svg figures
+mv figures/additional/national_r_3months.svg figures
+mv figures/additional/national_Rt_3months.svg figures
+
 git add data/*.csv
 git add outputs/*.csv
 git diff-index --quiet HEAD || git commit -m"Updated estimates $(date)"

### 2022-02-21 scripts/plot_estimates.r
tweak plots

@@ -87,7 +87,7 @@ non_variant <- data %>%
   filter(!grepl("^variant", level))
 
 histories <- list(all = Inf, `1year` = months(12), `3months` = months(3))
-breaks <- c(all = "2 months", `1year` = "1 month", `3months` = "1 month")
+breaks <- c(all = "4 months", `1year` = "3 month", `3months` = "1 month")
 
 for (history in names(histories)) {
   for (level in unique(non_variant$level)) {
@@ -113,7 +113,7 @@ for (history in names(histories)) {
       ggsave(here::here("figures", "additional",
 			paste0(level, "_", name, "_",
                                history, ".svg")), p,
-             width = 14, height = 8)
+             width = 10, height = 5)
     }
   }
 }

### 2022-02-21 scripts/update.sh
copy, not move

@@ -14,9 +14,9 @@ echo Variants
 git status | grep -q cis_variants.csv && Rscript scripts/estimate.R -v -d 1 && Rscript scripts/analyse.R -v
 
 Rscript scripts/plot_estimates.r
-mv figures/additional/national_infections_all.svg figures
-mv figures/additional/national_r_3months.svg figures
-mv figures/additional/national_Rt_3months.svg figures
+cp figures/additional/national_infections_all.svg figures
+cp figures/additional/national_r_3months.svg figures
+cp figures/additional/national_Rt_3months.svg figures
 
 git add data/*.csv
 git add outputs/*.csv

### 2022-02-21 scripts/simple-example.R
prettier table and move figures

@@ -78,7 +78,7 @@ prev_plot <- plot_prev(
   joint_data$prevalence[[1]]
 ) +
   scale_x_date(date_breaks = "4 months", date_label = "%b %Y")
-ggsave("figures/readme/prev.png", prev_plot, width = 9, height = 6)
+ggsave("figures/example/prev.png", prev_plot, width = 9, height = 6)
 
 # plot modelled and observed (but also modelled) antibodies
 ab_plot <- plot_prev(
@@ -90,7 +90,7 @@ ab_plot <- plot_prev(
   scale_y_continuous("Antibody prevalence", labels = scales::percent) +
   scale_x_date("Date", date_breaks = "4 months", date_label = "%b %Y")
 
-ggsave("figures/readme/ab.png", ab_plot, width = 9, height = 6)
+ggsave("figures/example/ab.png", ab_plot, width = 9, height = 6)
 
 # pairs plot
 stanfit <- read_stan_csv(fit$fit[[1]]$output_files())
@@ -99,7 +99,7 @@ pairs <- mcmc_pairs(fit$fit[[1]]$draws(),
   np = np,
   pars = c("beta[1]", "gamma[1]", "gamma[2]", "delta[1]", "k[1]", "l[1]")
 )
-ggsave("figures/readme/pairs.png", pairs, width = 16, height = 16)
+ggsave("figures/example/pairs.png", pairs, width = 16, height = 16)
 
 # plot infections
 inc_plot <- plot_trace(
@@ -108,7 +108,7 @@ inc_plot <- plot_trace(
   scale_y_continuous("Incident infections", labels = scales::percent) +
   scale_x_date("Date", date_breaks = "4 months", date_label = "%b %Y")
 
-ggsave("figures/readme/infections.png", inc_plot, width = 9, height = 6)
+ggsave("figures/example/infections.png", inc_plot, width = 9, height = 6)
 
 # plot growth
 growth_plot <- plot_trace(
@@ -116,7 +116,7 @@ growth_plot <- plot_trace(
 ) +
   labs(y = "Daily growth rate", x = "Date") +
   geom_hline(yintercept = 0, linetype = 2)
-ggsave("figures/readme/growth.png", growth_plot, width = 9, height = 6)
+ggsave("figures/example/growth.png", growth_plot, width = 9, height = 6)
 
 # plot Rt
 rt_plot <- plot_trace(
@@ -125,10 +125,10 @@ rt_plot <- plot_trace(
[…]

### 2022-02-21 docs/paper.Rmd
prettier table and move figures

--- 2022-02-17 writeup/paper.Rmd
@@
 Estimated reproduction numbers."}
-knitr::include_graphics(here::here("figures", "readme", "example-estimates.png"))
+knitr::include_graphics(here::here("figures", "example", "example-estimates.png"))
 ```
@@
   kbl(caption = "90% credible intervals (as quantiles of the posterior distribution) of biological parameters.",
-      booktabs = TRUE)
+      booktabs = TRUE) %>%
+  kable_styling()
 ```
@@
 ```{r pairs-plot, out.width=".7\\paperwidth", fig.cap = "Scatter plots and histograms of posterior parameter samples."}
-knitr::include_graphics(here::here("figures", "readme", "pairs.png"))
+knitr::include_graphics(here::here("figures", "example", "pairs.png"))
 ```

### 2022-02-21 docs/paper.Rmd
fix typo and update table

@@
            name == "beta", "Proportion infected that seroconvert",
-           name == "gamma (infection)", "Antibody waning following infection",
-           name == "gamma (vaccination)", "Antibody waning following vaccination",
+           name == "gamma (infection)", "Antibody waning following infection (daily rate)",
+           name == "gamma (vaccination)", "Antibody waning following vaccination (daily rate)",
            name == "delta", "Proportion vaccinated that seroconvert",
@@
 biol_params <- params[!is.na(Description)]
-biol_params[, `Estimate (90% CI)` := paste0(signif(q5, 2), "--", signif(q95, 2))]
+biol_params[, `Estimate (90% CI)` := paste0(signif(q50, 2), " (", signif(q5, 2), "--", signif(q95, 2), ")")]
 biol_params <- biol_params[, list(Parameter = name,
@@
 ```
+
+For correspondence contact: sebastian.funk@lshtm.ac.uk
 
@@
 biol_params %>%
-  kbl(caption = "90% credible intervals (as quantiles of the posterior distribution) of biological parameters.",
+  kbl(caption = "Estimates and 90% credible intervals (as quantiles of the posterior distribution) of biological parameters.",
       booktabs = TRUE) %>%
@@
 
-Posterior estimates of recovered biological parameters are shown in Table \@ref{tab:prams-table}.
+Posterior estimates of recovered biological parameters are shown in Table \@ref{tab:params-table}.
 Some of the parameter estimates show high levels of correlation suggesting issues of identifiability (Figure \@ref(fig:pairs-plot)).

### 2022-02-22 R/model.R
Antibody model fix (#27)
* swap antibody adjustment between infection/vaccination

* perhaps better: let the model decide scaling

* update prior

@@ -12,7 +12,7 @@ i2p_data <- function(prev, ab, vacc, init_ab,
                      prob_detectable, unobserved_time = 14, horizon = 0,
                      inf_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
                      vacc_ab_delay = c(rep(0, 7 * 4), rep(1 / 7, 7)),
-                     prop_dont_seroconvert = c(-2, 1), # 10%
+                     prop_seroconvert = c(2, 1), # 90%
                      inf_waning_rate = c(-9, 4),
                      vac_waning_rate = c(-9, 4), # 0.1%
                      vaccine_efficacy = c(3, 1), # 95%
@@ -123,7 +123,7 @@ i2p_data <- function(prev, ab, vacc, init_ab,
     prob_detect_sd = rev(prob_detectable$sd),
     pbt = max(prob_detectable$time) + 1,
     init_inc_mean = logit(init_inc_mean),
-    pbeta = prop_dont_seroconvert,
+    pbeta = prop_seroconvert,
     pgamma_mean = c(inf_waning_rate[1], vac_waning_rate[1]),
     pgamma_sd = c(inf_waning_rate[2], vac_waning_rate[2]),
     pdelta = vaccine_efficacy,

### 2022-02-23 scripts/estimate.R
don't produce local antibody estimates

@@ -119,7 +119,7 @@ if (antibodies) {
     data <- data %>%
       mutate(grouping = "all")
   } else if (local) {
-    ## estimate parameters from regions
+    stop("No local antibody estimates are available")
   } else if (age) {
     ## fit age groups without data alongside older ones
     data <- data %>%

### 2022-02-23 docs/paper.Rmd
fix typo
abstract

--- 2022-02-21 docs/paper.Rmd
@@
 ```
+
+For correspondence: sebastian.funk@lshtm.ac.uk
+
+# Abstract
+
+## Background
+
+Repeated measurements of cross-sectional prevalence of Polymerase Chain Reaction (PCR) positivity or seropositivity provide rich insight into the dynamics of an infection.
+The UK Office for National Statistics (ONS) Community Infection Survey publishes such measurements for SARS-CoV-2 on a weekly basis based on testing enrolled households, contributing to situational awareness in the country.
+Here we present estimates of time-varying and static epidemiological quantities that were derived from the estimates published by ONS.
+
+## Methods
+
+We used discrete convolutions based to estimate the incidence of infections based on previously published PCR detection curves describing the probability of a positive test as a function of the time since infection.
+We further combined time-varying estimates of antibody prevalence with a model of antibody positivity and waning that moved individuals between compartments with or without antibodies based on estimates of new infections, vaccination, probability of seroconversion and waning.
+
+## Results
+
+We produced incidence curves of infection describing the UK epidemic from late April 2020 until early 2022.
+We used these estimates of incidence to estimate the time-varying growth rate of infections, and combined them with estimates of the generation interval to estimate reproduction numbers.
+Biological parameters describing seroconversion and waning, while based on a simple model, were broadly in line with plausible ranges from individual-level studies.
+
+## Conclusions
+
+Beyond informing situational awareness and allowing for estimates using individual-level data, repeated cross-sectional studies make it possible to estimate epidemiological parameters from population-level models.
+Studies or public health surveillance methods based on similar designs offer opportunities for further improving our understanding of the dynamics of SARS-CoV-2 or other pathogens and their interaction with population-level immunity.
 

### 2022-02-23 docs/paper.Rmd
fixes

--- 2022-02-21 docs/paper.Rmd
@@
 
-For correspondence contact: sebastian.funk@lshtm.ac.uk
+For correspondence: sebastian.funk@lshtm.ac.uk
+
+# Abstract
+
+## Background
+
+Repeated measurements of cross-sectional prevalence of Polymerase Chain Reaction (PCR) positivity or seropositivity provide rich insight into the dynamics of an infection.
+The UK Office for National Statistics (ONS) Community Infection Survey publishes such measurements for SARS-CoV-2 on a weekly basis based on testing enrolled households, contributing to situational awareness in the country.
+Here we present estimates of time-varying and static epidemiological quantities that were derived from the estimates published by ONS.
+
+## Methods
+
+We used discrete convolutions based to estimate the incidence of infections based on previously published PCR detection curves describing the probability of a positive test as a function of the time since infection.
+We further combined time-varying estimates of antibody prevalence with a model of antibody positivity and waning that moved individuals between compartments with or without antibodies based on estimates of new infections, vaccination, probability of seroconversion and waning.
+
+## Results
+
+We produced incidence curves of infection describing the UK epidemic from late April 2020 until early 2022.
+We used these estimates of incidence to estimate the time-varying growth rate of infections, and combined them with estimates of the generation interval to estimate reproduction numbers.
+Biological parameters describing seroconversion and waning, while based on a simple model, were broadly in line with plausible ranges from individual-level studies.
+
+## Conclusions
+
+Beyond informing situational awareness and allowing for estimates using individual-level data, repeated cross-sectional studies make it possible to estimate epidemiological parameters from population-level models.
+Studies or public health surveillance methods based on similar designs offer opportunities for further improving our understanding of the dynamics of SARS-CoV-2 or other pathogens and their interaction with population-level immunity.
 

### 2022-02-23 docs/paper.Rmd
add interpretation of biological parameters

@@
     toc: false
+    number_sections: false
   bookdown::html_document2: default
@@
 Studies or public health surveillance methods based on similar designs offer opportunities for further improving our understanding of the dynamics of SARS-CoV-2 or other pathogens and their interaction with population-level immunity.
+
+\clearpage
 
@@
 ```{r params-table}
-biol_params %>%
-  kbl(caption = "Estimates and 90% credible intervals (as quantiles of the posterior distribution) of biological parameters.",
-      booktabs = TRUE) %>%
-  kable_styling()
+biol_params %>%  kable()
+  ## kable(caption = "Estimates and 90% credible intervals (CIs, as quantiles of the posterior distribution) of biological parameters.")
+  ##       booktabs = TRUE) %>%
+  ## kable_styling()
 ```
@@
 
+Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022].
+We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, compared to 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020].
+We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
+Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (infection)"]$q95, 2)`--`r signif(params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.
+
 As currently implemented, our method suffers from a number of limitations that risk biasing the results.
-Several of the key parameters in our model, especially the estimates of PCR positivity over time from infection and generation interval distributions, are fixed and based on estimates derived from wildtype virus in a particular cohort of healthcare workers and may well be incorrect for other circulating variants or populations.
+Several of the key parameters in our model, especially the estimates of PCR positivity over time from infection, generation interval distributions, are fixed and based on estimates derived from wildtype virus in a particular cohort of healthcare workers and may well be incorrect for other circulating variants or populations.
 Furthermore, generation times have been shown to change over time due to behavioural changes and epidemiological dynamics, which would affect our reproduction number estimates [@champredon2015; @hart2021; @park2021].
@@
 We modelled the growth of infections as a stationary Gaussian process, whereas in reality variation over time has changed between periods of stability and rapid change due to changes in contact behaviour in response to the epidemic.
-Lastly, we assumed that antibody waning was exponential and conferred either full protection if with detectable antibodies, or none whatsoever if not, and ignored any consequences of multiple rounds of vaccination or infection apart from converting those without detectable antibodies to having detectable antibodies.
+Lastly, we assumed that antibody waning was exponential, and ignored any consequences of multiple rounds of vaccination or infection apart from converting those without detectable antibodies to having detectable antibodies.
 

### 2022-02-24 R/postprocess.R
give index to gen_dab

@@ -62,7 +62,8 @@ i2p_add_var <- function(dt, prev, data, var_col = NULL) {
       ,
       paste(var_col) := fcase(
         name %in% c("est_prev", "est_ab", "infections",
-		    "dcases", "dab", "r", "R"), vars[n_index],
+		    "dcases", "dab", "gen_dab", "r", "R"), 
+			      vars[n_index],
         name %in% c("beta", "gamma", "delta", "k", "l"),
         paste0(vars[ab_index], collapse = ";")
       )

### 2022-02-24 docs/paper.Rmd
Draft writeup (#28)
* first draft of writeup

* add authors

* example fixes

* fix figure

* move to docs

* prettier table and move figures

* fix typo and update table

* first draft of writeup

* add authors

* example fixes

* fix figure

* move to docs

* prettier table and move figures

* fix typo

* abstract

* fixes

* add interpretation of biological parameters

* some more references and minor changes

@@
 which implies that growth, rather than incidence, remains at an estimated mean level in the absence of data, usually leading to better real-time performance [@abbott2020].
-The results shown in this paper were obtained using this formulation with a GP prior on the growth rate.
+The results shown here were obtained using this formulation with a GP prior on the growth rate.
 
@@
 
-When additionally using antibodies we convolve the modelled infections $I(t)$ as well as input data on vaccinations $Y_{\text{V}, t}$ with distributions quantifying the delay to generating detectable antibodies following infection (by default set to 4 weeks for both infection and vaccination), yielding potentially antibody-generating time series from infection $I^{\text{A}}$ and $V^{\text{A}}$.
+When additionally using antibodies we convolve the modelled infections $I(t)$ as well as input data on vaccinations $Y_{\text{V}, t}$ with distributions quantifying the delay to generating detectable antibodies following infection (by default set to 4 weeks for both infection and vaccination), yielding potentially antibody-generating time series from infection $I_{\text{A}}$ and $V_{\text{A}}$.
 We then calculate antibodies from infection as
 $$
-A^{\text{I}}(t) =
-A^{\text{I}}(t - 1) +
-\beta I^{\text{A}}(t)  (1 - A(t - 1))^k -
-\gamma_\text{I} A^{\text{I}}(t - 1)
+A_{\text{I}}(t) =
+A_{\text{I}}(t - 1) +
+\beta I_{\text{A}}(t)  (1 - A(t - 1))^k -
+\gamma_\text{I} A_{\text{I}}(t - 1)
 $$
@@
 $$
-A^{\text{V}}(t) =
-A^{\text{V}}(t - 1) +
-\delta V^{\text{A}}(t) (1 - A(t - 1))^l -
- \gamma_\text{V} A^{\text{V}}(t - 1)
+A_{\text{V}}(t) =
+A_{\text{V}}(t - 1) +
+\delta V_{\text{A}}(t) (1 - A(t - 1))^l -
+ \gamma_\text{V} A_{\text{V}}(t - 1)
 $$
@@
 $$
-A(t) = A^{\text{I}}(t)  + A^{\text{V}}(t)
+A(t) = A_{\text{I}}(t)  + A_{\text{V}}(t)
 $$
@@
 
-The model was implemented in _stan_ and using the _cmdstanr_ R package.
-All code needed to reproduce the results shown in this paper is available at https://github.com/epiforecasts/inc2prev.
+The model was implemented in _Stan_ and using the _cmdstanr_ R package [@gabry2021;@standev2022].
+All code needed to reproduce the results shown here is available at https://github.com/epiforecasts/inc2prev.
 
@@
 ```{r params-table}
-biol_params %>%  kable()
-  ## kable(caption = "Estimates and 90% credible intervals (CIs, as quantiles of the posterior distribution) of biological parameters.")
-  ##       booktabs = TRUE) %>%
-  ## kable_styling()
+biol_params %>% 
+  kbl(caption = paste0(
+    "Estimates and credible intervals (CIs, as quantiles of the posterior",
+    "distribution) of biological parameters."
+    ), booktabs = TRUE) %>%
+  kable_styling()
 ```
 
-Posterior estimates of recovered biological parameters are shown in Table \@ref{tab:params-table}.
+Posterior estimates of recovered biological parameters are shown in Table \@ref(tab:params-table).
 Some of the parameter estimates show high levels of correlation suggesting issues of identifiability (Figure \@ref(fig:pairs-plot)).
@@
 We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
-Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (infection)"]$q95, 2)`--`r signif(params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.
+Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (infection)"]$q95, 2)`--`r signif(1 / params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(1 / params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.
 
@@
 
-There is enormous potential for understanding epidemiological dynamics from repeated cross-sectional surveys.
+There is enormous potential for understanding epidemiological dynamics from repeated cross-sectional surveys, whether to identify current or past infection [@metcalf2016].
 Where the generation interval distribution is the same or close to the distribution of detectability after infection, this could be done using recently developed methods for unified modelling of incidence and prevalence [@pakkanen2021].
@@
 We thank Thomas House for insightful comments on this work, and the Office for National Statistics for making the data sets publicly available.
+We acknowledge funding by the Wellcome Trust (210758/Z/18/Z).
 

### 2022-02-24 R/model.R

@@ -203,6 +203,8 @@ i2p_inits <- function(dat) {
         beta = array(inv_logit(rnorm(1, -2, 0.4))),
         gamma = array(inv_logit(rnorm(2, -9, 0.4))),
         delta = array(inv_logit(rnorm(1, 3, 0.4))),
+        k = array(exp(rnorm(1, 0, 0.1))),
+        l = array(exp(rnorm(1, 0, 0.1))),
         ab_sigma = array(
           truncnorm::rtruncnorm(1, mean = 0.005, sd = 0.0025, a = 0)
         ),

### 2022-02-24 scripts/simple-example.R

@@ -15,10 +15,11 @@ library(tidyr)
 library(future.apply)
 library(future.callr)
 library(future)
+library(cowplot)
 
 # Test target
 example_var <- "England"
-end_date <- "2022-08-01"
+end_date <- "2021-11-01"
 ## Get tools
 functions <- list.files(here("R"), full.names = TRUE)
 walk(functions, source)
@@ -54,8 +55,12 @@ tune <- i2p_gp_tune_model()
 # Fit the infection to prevalence model
 fit <- incidence(
   joint_data$prevalence[[1]],
+  joint_data$antibodies[[1]],
+  joint_data$vaccination[[1]],
+  joint_data$initial_antibodies[[1]],
   variables = c(
-    "est_prev", "infections", "dcases", "r", "R"
+    "est_prev", "infections", "dcases", "r", "R",
+    "est_ab", "dab", "beta", "gamma", "delta", "k", "l"
   ),
   prob_detect = prob_detect, parallel_chains = 2, iter_warmup = 200,
   chains = 2, model = mod, adapt_delta = 0.85, max_treedepth = 15,
@@ -65,18 +70,15 @@ fit <- incidence(
   ),
   keep_fit = TRUE
 )
-fit
+fit$fit[[1]]$save_object(here::here("outputs", "example-fit.rds"))
 
 # plot modelled and observed (but also modelled) prevalence
 prev_plot <- plot_prev(
   fit$summary[[1]], fit$samples[[1]][sample <= 100],
   joint_data$prevalence[[1]]
-)
-ggsave("figures/prev.png", prev_plot, width = 9, height = 6)
-
-prev_logit_plot <- prev_plot +
-  scale_y_continuous(tran = scales::logit_trans())
-ggsave("figures/prev-logit.png", prev_logit_plot, width = 9, height = 6)
+) +
+  scale_x_date(date_breaks = "4 months", date_label = "%b %Y")
+ggsave("figures/example/prev.png", prev_plot, width = 9, height = 6)
 
 # plot modelled and observed (but also modelled) antibodies
 ab_plot <- plot_prev(
@@ -84,38 +86,53 @@ ab_plot <- plot_prev(
   joint_data$antibodies[[1]],
   data_source = "ONS Antibodies", observed = "est_ab",
   modelled = "dab"
-)
-ggsave("figures/ab.png", ab_plot, width = 9, height = 6)
+) +
[…]

### 2022-02-24 docs/report.Rmd
update README and add real-time report

---
title: "Real-time epidemiological estimates from ONS Community Infection Survey data"
author: Sam Abbott, Sebastian Funk
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output: html_document
---

```{r setup, echo = FALSE, include = FALSE}
library(cowplot)
library(here)

estimates <- c(national = "National",
               regional = "Regional",
               age_school = "Age")

quantities <- c(est_prev = "PCR positivity prevalence",
                infections = "Daily infection incidence",
                r = "Growth rate",
                Rt = "Reproduction number")

time_horizon <- c(est_prev = "1year",
                  infections = "1year",
                  r = "3months",
                  Rt = "3months")
```

Estimates were derived from the modelled daily and weekly data published weekly as part of the Covid-19 Community Infection Survey by the Office of National Statistics. Prevalence and incidence estimates are shown for the last year, and reproduction number for the past three months. Code to reproduce the results and links to data tables with estimates are available at https://github.com/epiforecasts/inc2prev

```{r, results = 'asis'}
for (level in names(estimates)) {
  cat("\n\n#", estimates[[level]], "estimates", "{.tabset}", "\n")
  for (quantity in names(quantities)) {
    cat("\n\n##", quantities[[quantity]], "\n")
    file_name <-
      paste0(level, "_", quantity, "_", time_horizon[[quantity]], ".svg")
    p <- ggdraw() + draw_image(here::here("figures", "additional", file_name))
    print(p)
  }
}
```

# License

This project uses data from the Office for National Statistics Community Infection Survey, which is licensed under the [Open Government License v3.0](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/datasets/coronaviruscovid19infectionsurveydata).

### 2022-02-24 R/model.R
add stan optimisation

@@ -243,9 +243,15 @@ i2p_inits <- function(dat) {
 #' @examplesIf interactive()
 #' mod <- i2p_model()
 i2p_model <- function(model = "stan/inc2prev.stan", include,
-                      compile = TRUE, threads = FALSE, verbose = TRUE, ...) {
+                      compile = TRUE, threads = FALSE, verbose = TRUE,
+                      optimise = TRUE, ...) {
   if (missing(include)) {
     include <- "stan/functions"
+  }
+  if (optimise) {
+    stanc_options <- list("O1")
+  } else {
+    stanc_options <- list()
   }
 
   if (compile) {
@@ -255,6 +261,7 @@ i2p_model <- function(model = "stan/inc2prev.stan", include,
         cpp_options = list(
           stan_threads = threads
         ),
+        stanc_options = stanc_options,
         ...
       )
     } else {
@@ -263,7 +270,9 @@ i2p_model <- function(model = "stan/inc2prev.stan", include,
           include_path = include,
           cpp_options = list(
             stan_threads = threads
-          ), ...
+          ),
+          stanc_options = stanc_options,
+	  ...
         )
       )
     }

### 2022-02-24 scripts/estimate.R
return all variables by default

@@ -158,14 +158,12 @@ incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, week
   )
   safe_incidence <- purrr::safely(incidence)
 
-  variables <- c("est_prev", "infections", "dcases", "r", "R")
   prev <- data$prevalence[[1]]
   if (weekly) {
     prev <- prev %>%
       convert_to_weekly(max(.$end_date))
   }
   if (antibodies) {
-    variables <- c(variables, "est_ab", "dab", "beta", "gamma", "delta")
     ab <- data$antibodies[[1]]
     vacc <- data$vaccination[[1]]
     init_ab <- data$initial_antibodies[[1]]
@@ -179,7 +177,6 @@ incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, week
     ab = ab,
     vacc = vacc,
     init_ab = init_ab,
-    variables = variables,
     prob_detect = pb, parallel_chains = 2, iter_warmup = 250,
     chains = 2, model = mod, adapt_delta = 0.9, max_treedepth = 12,
     data_args = list(

### 2022-02-24 scripts/estimate.R
remove superfluous model recompilation step

@@ -152,10 +152,6 @@ convert_to_weekly <- function(x, ref_end_date, cols = c("middle", "lower", "uppe
 incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, weekly = FALSE, ...) {
   message("Fitting model")
 
-  mod <- cmdstanr::cmdstan_model(
-    model$stan_file(),
-    include_paths = here::here("stan", "functions")
-  )
   safe_incidence <- purrr::safely(incidence)
 
   prev <- data$prevalence[[1]]
@@ -178,7 +174,7 @@ incidence_with_var <- function(data, pb, model, gp_model, differencing = 0, week
     vacc = vacc,
     init_ab = init_ab,
     prob_detect = pb, parallel_chains = 2, iter_warmup = 250,
-    chains = 2, model = mod, adapt_delta = 0.9, max_treedepth = 12,
+    chains = 2, model = model, adapt_delta = 0.9, max_treedepth = 12,
     data_args = list(
       gp_tune_model = gp_model, gp_m = gp_frac, differencing = differencing
     ),

### 2022-02-24 scripts/estimate.R
pass on weekly estimate

@@ -205,6 +205,7 @@ est <- future_lapply(
   pb = prob_detect,
   model = mod, gp_model = tune,
   differencing = differencing,
+  weekly = weekly,
   future.seed = TRUE
 )
 

### 2022-02-24 scripts/estimate.R
set variable names

@@ -219,6 +219,7 @@ est[, samples := map2(samples, level, ~ as.data.table(.x)[, level := .y])]
 # Bind posterior samples/summary together
 estimates <- bind_rows(est$summary)
 samples <- bind_rows(est$samples)
+est[, variable := data$variable]
 diagnostics <- select(est, -samples, -summary)
 
 suffix <- paste0(suffix, ifelse(antibodies, "_ab", ""))

### 2022-02-24 scripts/plot_estimates.r
tweak plots

@@ -23,7 +23,10 @@ files <-
 data <- map(files, read_csv) %>%
   bind_rows() %>%
   separate(variable, c("variable", "geography"), sep = "\\|")
-data <- left_join(data, local_region, by = c("level", "variable"))
+data <- left_join(data, local_region, by = c("level", "variable")) %>%
+  pivot_longer(starts_with("q"), names_to = "quantile") %>%
+  mutate(value = if_else(name == "infections", value / population, value)) %>%
+  pivot_wider(names_from = "quantile")
 
 cum_files <-
   list.files(here("outputs"), pattern = "cumulative_[^_]+\\.rds", full.names = TRUE)
@@ -63,16 +66,18 @@ var_names <- c(
   dcases = "Daily estimated prevalence",
   r = "Growth rate",
   Rt = "Reproduction number",
-  cumulative_infections = "Cumulative attack rate"
+  cumulative_infections = "Cumulative attack rate",
+  cumulative_exposure = "Cumulative exposure"
 )
 
 labels <- list(
   est_prev = scales::percent_format(1L, scale = 1),
-  infections = scales::comma,
+  infections = scales::percent_format(0.1),
   dcases = scales::percent_format(1L),
   r = waiver(),
   Rt = waiver(),
-  cumulative_infections = scales::percent_format(1L)
+  cumulative_infections = scales::percent_format(1L),
+  cumulative_exposure = scales::percent_format(1L)
 )
 
 group <- c(
@@ -81,6 +86,11 @@ group <- c(
   regional = "Region",
   variant_regional = "Variant",
   variant_national = "Variant"
+)
+
+hline <- c(
+  r = 0,
+  Rt = 1
 )
 
 non_variant <- data %>%
@@ -110,6 +120,9 @@ for (history in names(histories)) {
         scale_colour_brewer(group[level],  palette = "Set1") +
         scale_fill_brewer(group[level], palette = "Set1") +
         theme_minimal()
+      if (name %in% names(hline)) {
+        p <- p + geom_hline(yintercept = hline[[name]], linetype = "dashed")
+      }
       ggsave(here::here("figures", "additional",
 			paste0(level, "_", name, "_",
                                history, ".svg")), p,
@@ -146,6 +159,9 @@ for (history in names(histories)) {
[…]

### 2022-02-24 scripts/publish.sh
separate update and publication

@@ -1,0 +1,7 @@
+#!/bin/sh
+
+git pull -Xours
+git add data/*.csv
+git add outputs/*.csv
+git diff-index --quiet HEAD || git commit -m"Updated estimates $(date)"
+git push -v

### 2022-02-24 scripts/update.sh

@@ -5,7 +5,8 @@ git pull -Xours
 Rscript data-raw/update_cis.R
 
 echo Space
-git status | grep -q cis.csv && Rscript scripts/estimate.R -n -d 1 && Rscript scripts/analyse.R
+git status | grep -q cis.csv && Rscript scripts/estimate.R -d 1 && Rscript scripts/analyse.R
+git status | grep -q cis.csv && Rscript scripts/estimate.R -n -r -d 1 && Rscript scripts/analyse.R -r
 
 echo Age
 git status | grep -q cis_age.csv && Rscript scripts/estimate.R -g -d 1 && Rscript scripts/analyse.R -g
@@ -18,7 +19,3 @@ cp figures/additional/national_infections_all.svg figures
 cp figures/additional/national_r_3months.svg figures
 cp figures/additional/national_Rt_3months.svg figures
 
-git add data/*.csv
-git add outputs/*.csv
-git diff-index --quiet HEAD || git commit -m"Updated estimates $(date)"
-git push -v

### 2022-02-24 R/plot_wrapper.R
plot crude cumulative exposure

@@ -51,9 +51,12 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
       inner_join(early_samples, by = c("variable", "sample")) %>%
       mutate(value = value + initial) %>%
       select(-initial)
+    cumulative_exposure_samples <- combined_samples %>%
+      mutate(value = 1 - exp(-value),
+	     name = "cumulative_exposure")
     level_samples <- level_samples %>%
-      filter(name != "cumulative_infections") %>%
-      bind_rows(combined_samples)
+      filter(!(name %in% c("cumulative_infections", "cumulative_exposure"))) %>%
+      bind_rows(combined_samples, cumulative_exposure_samples)
   }
   ## split geography and variants
   if (grepl("^variant_", level)) {
@@ -102,7 +105,7 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
   ## 1) plot prevalence
   p <- plot_prev(level_estimates, level_samples, level_prev) +
     facet_wrap(plot_formula, scales = "free_y")
-  ggsave(here::here("figures", paste0("prev_", level, suffix, extension)), p,
+  ggsave(here::here("figures", "additional", paste0("prev_", level, suffix, extension)), p,
     width = 7 + 3 * floor(sqrt(nvars)),
     height = 2 + 3 * floor(sqrt(nvars))
   )
@@ -111,7 +114,7 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
       modelled = "dab", observed = "est_ab"
     ) +
       facet_wrap(plot_formula)
-    ggsave(here::here("figures", paste0("ab_", level, suffix, extension)), p,
+    ggsave(here::here("figures", "additional", paste0("ab_", level, suffix, extension)), p,
       width = 7 + 3 * floor(sqrt(nvars)),
       height = 2 + 3 * floor(sqrt(nvars))
     )
@@ -119,7 +122,7 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
   ## 2) plot incidence
   p <- plot_trace(level_samples, "infections") +
     facet_wrap(plot_formula, scales = "free_y")
-  ggsave(here::here("figures", paste0("inf_", level, suffix, extension)), p,
+  ggsave(here::here("figures", "additional", paste0("inf_", level, suffix, extension)), p,
[…]

### 2022-02-24 scripts/analyse.R

@@ -81,7 +81,8 @@ updated_samples <- map(
 )
 
 cumulative <- bind_rows(updated_samples) %>%
-  filter(name == "cumulative_infections")
+  filter(name %in% c("cumulative_infections",
+                     "cumulative_exposure"))
 
 saveRDS(cumulative,
         here::here("outputs", paste0("cumulative", suffix, ".rds")))

### 2022-02-24 scripts/simple-example.R
example: save combined plot and parameter estimates for writeup

@@ -65,8 +65,11 @@ fit <- incidence(
   ),
   keep_fit = TRUE
 )
-fit
+dir.create(here::here("outputs"), showWarnings = FALSE)
+fit$fit[[1]]$save_object(here::here("outputs", "example-fit.rds"))
 
+dir.create(here::here("figures", "example"), 
+	   showWarnings = FALSE, recursive = TRUE)
 # plot modelled and observed (but also modelled) prevalence
 prev_plot <- plot_prev(
   fit$summary[[1]], fit$samples[[1]][sample <= 100],
@@ -118,4 +121,14 @@ plot_trace(
 ) +
   labs(y = "Effective reproduction number", x = "Date") +
   geom_hline(yintercept = 1, linetype = 2)
-ggsave("figures/Rt.png", width = 9, height = 6)
+ggsave("figures/example/Rt.png", rt_plot, width = 9, height = 6)
+
+p <- plot_grid(prev_plot, inc_plot, ab_plot, rt_plot, labels = c("A", "B", "C", "D"))
+ggsave("figures/example/example-estimates.png", p, width = 12, height = 6)
+
+saveRDS(fit$summary[[1]], here::here("outputs", "example-summary.rds"))
+saveRDS(fit$samples[[1]], here::here("outputs", "example-samples.rds"))
+
+params <- fit$summary[[1]][is.na(date)]
+params[, c("n_index", "t_index", "date") := NULL]
+fwrite(params, "outputs/example-parameters.csv")

### 2022-02-24 scripts/simple-example.R
figures locations in example script

@@ -75,11 +75,11 @@ prev_plot <- plot_prev(
   fit$summary[[1]], fit$samples[[1]][sample <= 100],
   joint_data$prevalence[[1]]
 )
-ggsave("figures/prev.png", prev_plot, width = 9, height = 6)
+ggsave(here::here("figures", "prev.png"), prev_plot, width = 9, height = 6)
 
 prev_logit_plot <- prev_plot +
   scale_y_continuous(tran = scales::logit_trans())
-ggsave("figures/prev-logit.png", prev_logit_plot, width = 9, height = 6)
+ggsave(here::here("figures", "prev-logit.png"), prev_logit_plot, width = 9, height = 6)
 
 # plot modelled and observed (but also modelled) antibodies
 ab_plot <- plot_prev(
@@ -88,7 +88,7 @@ ab_plot <- plot_prev(
   data_source = "ONS Antibodies", observed = "est_ab",
   modelled = "dab"
 )
-ggsave("figures/ab.png", ab_plot, width = 9, height = 6)
+ggsave(here::here("figures", "ab.png"), ab_plot, width = 9, height = 6)
 
 # pairs plot
 stanfit <- read_stan_csv(fit$fit[[1]]$output_files())
@@ -97,7 +97,7 @@ pairs <- mcmc_pairs(fit$fit[[1]]$draws(),
   np = np,
   pars = c("alpha", "rho", "beta", "gamma[1]", "gamma[2]", "delta")
 )
-ggsave("figures/pairs.png", pairs, width = 16, height = 16)
+ggsave(here::here("figures", "pairs.png"), pairs, width = 16, height = 16)
 
 # plot infections
 plot_trace(
@@ -105,15 +105,14 @@ plot_trace(
 ) +
   labs(y = "Infections", x = "Date") +
   scale_y_continuous(labels = scales::percent)
-ggsave("figures/infections.png", width = 9, height = 6)
 
 # plot growth
-plot_trace(
+r_plot <- plot_trace(
   fit$samples[[1]][sample <= 100], "r"
 ) +
   labs(y = "Daily growth rate", x = "Date") +
   geom_hline(yintercept = 0, linetype = 2)
-ggsave("figures/growth.png", width = 9, height = 6)
+ggsave(here::here("figures", "growth.png"), width = 9, height = 6)
 
 # plot Rt
 plot_trace(
@@ -121,10 +120,10 @@ plot_trace(
 ) +
   labs(y = "Effective reproduction number", x = "Date") +
[…]

### 2022-02-24 docs/report.Rmd
remove code block

@@
 Prevalence and incidence estimates are shown for the last year, and reproduction number for the past three months.
-Code to reproduce the results and links to data tables with estimates are available at https://github.com/epiforecasts/inc2prev
+Code to reproduce the results and links to data tables with estimates are available at [https://github.com/epiforecasts/inc2prev](https://github.com/epiforecasts/inc2prev).
 
-```{r, results = 'asis'}
+```{r, results = 'asis', echo = FALSE}
 for (level in names(estimates)) {

### 2022-02-24 scripts/publish.sh
add report publication to update

@@ -5,3 +5,12 @@ git add data/*.csv
 git add outputs/*.csv
 git diff-index --quiet HEAD || git commit -m"Updated estimates $(date)"
 git push -v
+
+git branch -d gh-pages
+git checkout --orphan gh-pages
+git rm -rf .
+git add docs/report.html
+git commit -m "Update report"
+git push --force origin gh-pages
+git checkout main
+

### 2022-02-24 scripts/update.sh

@@ -15,7 +15,4 @@ echo Variants
 git status | grep -q cis_variants.csv && Rscript scripts/estimate.R -v -d 1 && Rscript scripts/analyse.R -v
 
 Rscript scripts/plot_estimates.r
-cp figures/additional/national_infections_all.svg figures
-cp figures/additional/national_r_3months.svg figures
-cp figures/additional/national_Rt_3months.svg figures
-
+Rscript -e 'rmarkdown::render("docs/report.Rmd")'

### 2022-02-25 scripts/publish.sh
update scripts

@@ -6,10 +6,10 @@ git add outputs/*.csv
 git diff-index --quiet HEAD || git commit -m"Updated estimates $(date)"
 git push -v
 
-git branch -d gh-pages
+git branch -D gh-pages
 git checkout --orphan gh-pages
 git rm -rf .
-git add docs/report.html
+git add docs/*.html
 git commit -m "Update report"
 git push --force origin gh-pages
 git checkout main

### 2022-02-25 scripts/update.sh

@@ -18,3 +18,4 @@ git status | grep -q cis_variants.csv && Rscript scripts/estimate.R -v -d 1 && R
 
 Rscript scripts/plot_estimates.r
 Rscript -e 'rmarkdown::render("docs/report.Rmd")'
+Rscript -e 'rmarkdown::render("docs/paper.Rmd")'

### 2022-02-28 scripts/plot_estimates.r
more plots

@@ -7,6 +7,10 @@ library(readr)
 library(scales)
 library(tidyr)
 
+## whether to make spaghetti plots
+spaghetti <- FALSE
+
+## Get tools
 source(here::here("R", "read.R"))
 prev <- read_cis()
 local_region <- prev %>%
@@ -16,49 +20,6 @@ local_region <- prev %>%
 
 dir.create(here::here("figures", "additional"),
            showWarnings = FALSE, recursive = TRUE)
-
-## Get tools
-files <-
-  list.files(here("outputs"), pattern = "estimates_[^_]+\\.csv", full.names = TRUE)
-data <- map(files, read_csv) %>%
-  bind_rows() %>%
-  separate(variable, c("variable", "geography"), sep = "\\|")
-data <- left_join(data, local_region, by = c("level", "variable")) %>%
-  pivot_longer(starts_with("q"), names_to = "quantile") %>%
-  mutate(value = if_else(name == "infections", value / population, value)) %>%
-  pivot_wider(names_from = "quantile")
-
-cum_files <-
-  list.files(here("outputs"), pattern = "cumulative_[^_]+\\.rds", full.names = TRUE)
-if (length(cum_files) > 0) {
-  cum_data <- map(cum_files, readRDS) %>%
-    bind_rows() %>%
-    group_by(name, date, variable, variant, level) %>%
-    summarise(
-      q5 = quantile(value, 0.05),
-      q95 = quantile(value, 0.95),
-      q25 = quantile(value, 0.25),
-      q75 = quantile(value, 0.75),
-      q45 = quantile(value, 0.45),
-      q55 = quantile(value, 0.55),
-      .groups = "drop"
-    ) %>%
-    mutate(geography = if_else(
-             !is.na(variant),
-             as.character(variable),
-             NA_character_
-           ),
-           variable = if_else(
-             is.na(variant),
-             as.character(variable),
-             variant))
-  data <- data %>%
-    bind_rows(cum_data)
-}
-
-data <- data %>%
-  mutate(variable = if_else(variable == "2-10", "02-10", variable),
-         name = if_else(name == "R", "Rt", name))  
 
 var_names <- c(
   est_prev = "Prevalence estimate",
@@ -93,80 +54,107 @@ hline <- c(
   Rt = 1
 )
 
-non_variant <- data %>%
-  filter(!grepl("^variant", level))
-
 histories <- list(all = Inf, `1year` = months(12), `3months` = months(3))
[…]

### 2022-02-28 scripts/publish.sh
correct branch

@@ -12,5 +12,5 @@ git rm -rf .
 git add docs/*.html
 git commit -m "Update report"
 git push --force origin gh-pages
-git checkout main
+git checkout master
 

### 2022-03-01 docs/report.Rmd
load magick library
use png in report

@@
 library(here)
+library(magick)
 
@@
     file_name <-
-      paste0(level, "_", quantity, "_", time_horizon[[quantity]], ".svg")
+      paste0(level, "_", quantity, "_", time_horizon[[quantity]], ".png")
     p <- ggdraw() + draw_image(here::here("figures", "additional", file_name))

### 2022-03-01 scripts/plot_estimates.r
tweak plots

@@ -117,11 +117,13 @@ for (file in files) {
         plot_df <- level_data %>%
           filter(name == {{ name }}, 
 		 date > max(date) - histories[[history]])
-        aes_str <- list(x = "date", colour = colour_var)
+        aes_str <- list(x = "date")
         if (spaghetti) {
           plot_df <- plot_df %>%
             mutate(var_sample = interaction(variable, sample))
-          aes_str <- c(aes_str, list(y = "value", group = "var_sample"))
+          aes_str <- c(
+	    aes_str, list(y = "value", group = "var_sample", colour = colour_var)
+	  )
         } else {
           aes_str <- c(aes_str, list(fill = colour_var))
         }
@@ -132,8 +134,9 @@ for (file in files) {
             geom_line(alpha = 0.25)
         } else {
           p <- p +
-            geom_ribbon(mapping = aes(ymin = `q45`, ymax = `q55`), alpha = 0.5) +
-            geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.125)
+            geom_ribbon(mapping = aes(ymin = `q45`, ymax = `q55`), alpha = 0.5, colour = NA) +
+            geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.25, colour = NA)
+            geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.1, colour = NA)
         }
         p <- p +
           scale_x_date(breaks = breaks[[history]],

### 2022-03-01 scripts/plot_estimates.r
optional filetype

@@ -9,6 +9,7 @@ library(tidyr)
 
 ## whether to make spaghetti plots
 spaghetti <- FALSE
+filetype <- "png"
 
 ## Get tools
 source(here::here("R", "read.R"))
@@ -155,7 +156,7 @@ for (file in files) {
         }
         ggsave(here::here("figures", "additional",
                           paste0(level, "_", name, "_",
-                                 history, ".svg")), p,
+                                 history, ".", filetype)), p,
                width = 10, height = 5)
       }
     }

### 2022-03-01 scripts/plot_estimates.r
add colour

@@ -118,13 +118,11 @@ for (file in files) {
         plot_df <- level_data %>%
           filter(name == {{ name }}, 
 		 date > max(date) - histories[[history]])
-        aes_str <- list(x = "date")
+        aes_str <- list(x = "date", colour = colour_var)
         if (spaghetti) {
           plot_df <- plot_df %>%
             mutate(var_sample = interaction(variable, sample))
-          aes_str <- c(
-	    aes_str, list(y = "value", group = "var_sample", colour = colour_var)
-	  )
+          aes_str <- c(aes_str, list(y = "value", group = "var_sample"))
         } else {
           aes_str <- c(aes_str, list(fill = colour_var))
         }
@@ -135,9 +133,9 @@ for (file in files) {
             geom_line(alpha = 0.25)
         } else {
           p <- p +
-            geom_ribbon(mapping = aes(ymin = `q45`, ymax = `q55`), alpha = 0.5, colour = NA) +
-            geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.25, colour = NA)
-            geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.1, colour = NA)
+            geom_ribbon(mapping = aes(ymin = `q45`, ymax = `q55`)) +
+            geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.35, colour = NA)
+            geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.175, colour = NA)
         }
         p <- p +
           scale_x_date(breaks = breaks[[history]],
@@ -145,7 +143,7 @@ for (file in files) {
           scale_y_continuous(var_names[name], labels = labels[[name]]) +
           scale_colour_brewer(group[level],  palette = "Set1") +
           scale_fill_brewer(group[level],  palette = "Set1") +
-          theme_minimal()
+          theme_light()
         if (name %in% names(hline)) {
           p <- p +
             geom_hline(yintercept = hline[[name]], linetype = "dashed")

### 2022-03-01 scripts/update.sh
all formats

@@ -18,4 +18,4 @@ git status | grep -q cis_variants.csv && Rscript scripts/estimate.R -v -d 1 && R
 
 Rscript scripts/plot_estimates.r
 Rscript -e 'rmarkdown::render("docs/report.Rmd")'
-Rscript -e 'rmarkdown::render("docs/paper.Rmd")'
+Rscript -e 'rmarkdown::render("docs/paper.Rmd", output_format = "all")'

### 2022-03-01 scripts/plot_estimates.r
add line

@@ -33,7 +33,7 @@ var_names <- c(
 )
 
 labels <- list(
-  est_prev = scales::percent_format(1L, scale = 1),
+  est_prev = scales::percent_format(0.1, scale = 1),
   infections = scales::percent_format(0.1),
   dcases = scales::percent_format(1L),
   r = waiver(),
@@ -124,7 +124,7 @@ for (file in files) {
             mutate(var_sample = interaction(variable, sample))
           aes_str <- c(aes_str, list(y = "value", group = "var_sample"))
         } else {
-          aes_str <- c(aes_str, list(fill = colour_var))
+          aes_str <- c(aes_str, list(y = "q50", fill = colour_var))
         }
         p <- ggplot(plot_df, do.call(aes_string, aes_str)) +
           ylab(name) + xlab("")
@@ -133,7 +133,7 @@ for (file in files) {
             geom_line(alpha = 0.25)
         } else {
           p <- p +
-            geom_ribbon(mapping = aes(ymin = `q45`, ymax = `q55`)) +
+            geom_line()
             geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.35, colour = NA)
             geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.175, colour = NA)
         }
@@ -155,7 +155,7 @@ for (file in files) {
         ggsave(here::here("figures", "additional",
                           paste0(level, "_", name, "_",
                                  history, ".", filetype)), p,
-               width = 10, height = 5)
+               width = 8, height = 4)
       }
     }
   }

### 2022-03-01 R/plot_wrapper.R
always estimate exposure

@@ -51,13 +51,16 @@ plot_wrapper <- function(level, prev, ab = NULL, samples, estimates, early = NUL
       inner_join(early_samples, by = c("variable", "sample")) %>%
       mutate(value = value + initial) %>%
       select(-initial)
-    cumulative_exposure_samples <- combined_samples %>%
-      mutate(value = 1 - exp(-value),
-	     name = "cumulative_exposure")
     level_samples <- level_samples %>%
-      filter(!(name %in% c("cumulative_infections", "cumulative_exposure"))) %>%
-      bind_rows(combined_samples, cumulative_exposure_samples)
+      filter(!(name == c("cumulative_infections"))) %>%
+      bind_rows(combined_samples)
   }
+  cumulative_exposure_samples <- level_samples %>%
+    filter(name == "cumulative_infections") %>%
+    mutate(value = 1 - exp(-value),
+           name = "cumulative_exposure")
+  level_samples <- level_samples %>%
+    bind_rows(cumulative_exposure_samples)
   ## split geography and variants
   if (grepl("^variant_", level)) {
     level_prev <- level_prev %>%

### 2022-03-02 scripts/plot_estimates.r
add missing pluses

@@ -133,8 +133,8 @@ for (file in files) {
             geom_line(alpha = 0.25)
         } else {
           p <- p +
-            geom_line()
-            geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.35, colour = NA)
+            geom_line() +
+            geom_ribbon(mapping = aes(ymin = `q25`, ymax = `q75`), alpha = 0.35, colour = NA) +
             geom_ribbon(mapping = aes(ymin = `q5`, ymax = `q95`), alpha = 0.175, colour = NA)
         }
         p <- p +

### 2022-03-02 scripts/plot_estimates.r
reword

@@ -28,7 +28,7 @@ var_names <- c(
   dcases = "Daily estimated prevalence",
   r = "Growth rate",
   Rt = "Reproduction number",
-  cumulative_infections = "Cumulative attack rate",
+  cumulative_infections = "Cumulative infections",
   cumulative_exposure = "Cumulative exposure"
 )
 

### 2022-03-02 docs/report.Rmd
additional text

@@
 
-quantities <- c(est_prev = "PCR positivity prevalence",
-                infections = "Daily infection incidence",
+quantities <- c(est_prev = "PCR+ prevalence",
+                infections = "Infection incidence",
                 r = "Growth rate",
-                Rt = "Reproduction number")
+                Rt = "Reproduction number",
+                cumulative_infections = "Cumulative incidence",
+                cumulative_exposure = "Proportion ever infected")
 
@@
                   Rt = "3months")
+
+additional_text <- list(
+  est_prev = "Proportion testing PCR positive.",
+  infections = "Daily proportion newly infected.",
+  r = "Growth rate of new infections.",
+  Rt = "Infection reproduction number.",
+  cumulative_infections = "Cumulative incidence of infections, including reinfections.",
+  cumulative_exposure = "Crude estimate of the number of people ever infected under the extremely simplifying assumption that the risk of infection is independent of prior infection or vaccination status."
+)
 ```
@@
     cat("\n\n##", quantities[[quantity]], "\n")
+    cat(additional_text[[quantity]], "\n")
     file_name <-

### 2022-03-02 scripts/plot_estimates.r
adapt scales/labels

@@ -28,12 +28,12 @@ var_names <- c(
   dcases = "Daily estimated prevalence",
   r = "Growth rate",
   Rt = "Reproduction number",
-  cumulative_infections = "Cumulative infections",
-  cumulative_exposure = "Cumulative exposure"
+  cumulative_infections = "Cumulative incidence",
+  cumulative_exposure = "Proportion ever infected"
 )
 
 labels <- list(
-  est_prev = scales::percent_format(0.1, scale = 1),
+  est_prev = scales::percent_format(1L),
   infections = scales::percent_format(0.1),
   dcases = scales::percent_format(1L),
   r = waiver(),

### 2022-03-02 scripts/plot_estimates.r
expand plots

@@ -152,6 +152,10 @@ for (file in files) {
           p <- p +
             facet_wrap(~geography)
         }
+	if (grepl("cumulative_", name)) {
+	  p <- p + 
+            expand_limits(y = 0)
+	}
         ggsave(here::here("figures", "additional",
                           paste0(level, "_", name, "_",
                                  history, ".", filetype)), p,

### 2022-03-02 docs/report.Rmd
add horizons
text tweaks

@@
                 r = "Growth rate",
-                Rt = "Reproduction number",
+                Rt = "R",
                 cumulative_infections = "Cumulative incidence",
@@
                   r = "3months",
-                  Rt = "3months")
+                  Rt = "3months",
+                  cumulative_infections = "1year",
+                  cumulative_exposure = "1year")
 
@@
     cat("\n\n##", quantities[[quantity]], "\n")
-    cat(additional_text[[quantity]], "\n")
+    cat(additional_text[[quantity]], "\n\n")
     file_name <-

### 2022-03-02 docs/report.Rmd
add links

@@
 
-Estimates were derived from the modelled daily and weekly data published weekly as part of the Covid-19 Community Infection Survey by the Office of National Statistics.
+Estimates were derived from the modelled daily and weekly data published weekly as part of the [Coronavirus (COVID-19) Infection Survey](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/previousReleases) by the [Office of National Statistics](https://www.ons.gov.uk/).
 Prevalence and incidence estimates are shown for the last year, and reproduction number for the past three months.

### 2022-03-06 docs/paper.Rmd
read data read in

@@
 params <- fread(params_table)
-params[name == "gamma", name := ifelse(index == 1, "gamma (infection)", "gamma (vaccination)")]
-params[, Description := fcase(
+params[name == "gamma",
+       name := ifelse(index == 1, "gamma (infection)", "gamma (vaccination)")
+]
+params[,
+  Description := fcase(
     name == "beta", "Proportion infected that seroconvert",
@@
     name == "k", "Efficacy adjustment of immunity following infection",
-           name == "l", "Efficacy adjustment of immunity following vaccination")]
+    name == "l", "Efficacy adjustment of immunity following vaccination")
+]
 biol_params <- params[!is.na(Description)]
-biol_params[, `Estimate (90% CI)` := paste0(signif(q50, 2), " (", signif(q5, 2), "--", signif(q95, 2), ")")]
-biol_params <- biol_params[, list(Parameter = name,
-                                  Description,
-                                  `Estimate (90% CI)`)]
+biol_params[, 
+  `Estimate (90% CI)` := paste0(
+    signif(q50, 2), " (", signif(q5, 2), "--", signif(q95, 2), ")"
+  )
+]
+biol_params <- biol_params[,
+  .(Parameter = name, Description, `Estimate (90% CI)`)
+]
 

### 2022-03-06 scripts/simple-example.R
use data.table, build pkg docs, switch to using load_all() everywhere

@@ -21,8 +21,7 @@ library(cowplot)
 example_var <- "England"
 end_date <- "2021-11-01"
 ## Get tools
-functions <- list.files(here("R"), full.names = TRUE)
-walk(functions, source)
+devtools::load_all()
 
 # Load prevalence data and split by location
 prev <- read_cis() %>%

### 2022-03-06 docs/paper.Rmd
use data.table, build pkg docs, switch to using load_all() everywhere
abstract

@@
 ## Get tools
-functions <- list.files(here("R"), full.names = TRUE)
-walk(functions, source)
+devtools::load_all()
 
@@
 
-We used discrete convolutions based to estimate the incidence of infections based on previously published PCR detection curves describing the probability of a positive test as a function of the time since infection.
-We further combined time-varying estimates of antibody prevalence with a model of antibody positivity and waning that moved individuals between compartments with or without antibodies based on estimates of new infections, vaccination, probability of seroconversion and waning.
+We used a guassian process to model incidence of infections and then estimated observed PCR prevalence by convolving our modelled incidence estimates with a previously published PCR detection curve describing the probability of a positive test as a function of the time since infection.
+We refined our incidence estimates using time-varying estimates of antibody prevalence combined with a model of antibody positivity and waning that moved individuals between compartments with or without antibodies based on estimates of new infections, vaccination, probability of seroconversion and waning.
 
@@
 We produced incidence curves of infection describing the UK epidemic from late April 2020 until early 2022.
-We used these estimates of incidence to estimate the time-varying growth rate of infections, and combined them with estimates of the generation interval to estimate reproduction numbers.
+We used these estimates of incidence to estimate the time-varying growth rate of infections, and combined them with estimates of the generation interval to estimate time-varying reproduction numbers.
 Biological parameters describing seroconversion and waning, while based on a simple model, were broadly in line with plausible ranges from individual-level studies.

### 2022-03-06 docs/paper.Rmd
introduction

@@
 While repeated randomised cross-sectional sampling of positivity and antibodies provides utility in themselves for tracking an epidemic in real time, they can also be used for estimating epidemiological quantities by combining them with information on infection kinetics and immunological responses.
-Here we present a semi-mechanistic model that combines PCR positivity curves, generation interval estimates and vaccination data with ONS PCR positivity and antibody data to estimate infection incidence and its growth rate, reproduction numbers and rates of antibody waning.
+Here we present a semi-mechanistic model that combines PCR positivity curves, generation interval estimates and vaccination data with ONS PCR positivity and antibody data to estimate infection incidence and its growth rates, reproduction numbers and rates of antibody waning.
 

### 2022-03-06 docs/paper.Rmd
methods

@@
 
-We obtained the published estimates of daily prevalence of Polymerase Chain Reaction (PCR) positivity beginning on 26 April, 2020, from the ONS Community infection survey separately by nation, region, age group and variant, alongside their 95% confidence intervals, from the published spreadsheets on the ONS web site.
+We obtained the published estimates of daily prevalence of Polymerase Chain Reaction (PCR) positivity beginning on 26 April, 2020, from the ONS Community infection survey separately by nation, region, age group and variant, alongside their 95% credible intervals, from the published spreadsheets on the ONS web site.
 ONS estimates of a given prevalence vary between publication dates as the internal model to calculate prevalence involves smoothing, such that new data points in the present affect the estimates of times past.
-We aggregated estimates of PCR positivity for a single day produced for different publication dates by calculating the central estimate and confidence limits as the medians of the different respective central estimates and confidence limits.
+We aggregated estimates of PCR positivity for a single day produced for different publication dates by calculating the central estimate and credible intervals as the medians of the different respective central estimates and credible intervals.
 
@@
 
-We developed Bayesian model to estimate epidemiological quantities from ONS PCR positivity estimates and, optionally, population level antibody prevalence estimates and vaccination coverage.
+We developed a Bayesian model to estimate epidemiological quantities from ONS PCR positivity estimates and, optionally, population level antibody prevalence estimates and vaccination coverage.
 
@@
 
-We estimated the population proportion newly infected in the population $I(t)$ as a latent variable that is convolved with an PCR positivity curve $p(s)$, the probability of someone infected at time $s=0$ to test PCR positive to yield prevalence of PCR positivity $P(t)$.
+We estimated the population proportion newly infected in the population $I(t)$ as a latent variable that is convolved with a PCR positivity curve $p(s)$, the probability of someone infected at time $s=0$ to test PCR positive to yield prevalence of PCR positivity $P(t)$.
 $$
@@
 where $t_\text{p,max}=60$ is the maximum time modelled for which a person can stay PCR positive.
-We assumed each $p(s)$ to have an independent normal prior distribution at each time $s$ after infection with given mean and standard deviation estimates from another study [@hellewell2021].
-Infection incidence $I(t)$ is distinct form the estimates of PCR positivity incidence provided by ONS alongside the prevalence estimates, as it allows for the probability of infections testing yielding negative PCR results as a function of the time since infection.
+We assumed each $p(s)$ to have an independent normal prior distribution at each time $s$ after infection with given mean and standard deviation estimates from the posterior estimates of another study [@hellewell2021].
+Infection incidence $I(t)$ is distinct from the estimates of PCR positivity incidence provided by ONS alongside the prevalence estimates, as it allows for the probability of infections yielding negative PCR results as a function of the time since infection and is indexed by date of infection rather than the date of first testing positive.
 
 We used Gaussian Process (GP) priors to ensure smoothness of the estimates and deal with data gaps, whereby alternatively either $I(t)$ is has a GP prior with exponential quadratic kernel.
+To reduce the computational requirements of our approach we used an approximate rather than exact GP [@approxGP].
 $$
 \begin{aligned}
-I(t) &\sim \text{logit}(i_0 + i(t))\\
+\text{logit} \left( I(t) \right) &\sim i_0 + i(t)\\
 i(t) &\sim \text{GP}(t)
@@
 $$
-where $i_0$ is the estimated mean of the GP, or the GP prior is applied to higher order differences, for example the growth rate such as
+where $i_0$ is the estimated mean of the GP, or the GP prior is applied to higher order differences when infections are non-stationary, for example the growth rate such as
 $$
@@
 $$
-which implies that growth, rather than incidence, remains at an estimated mean level in the absence of data, usually leading to better real-time performance [@abbott2020].
+which implies that growth tends to zero when outside the range of the data, usually leading to better real-time performance [@abbott2020].
 The results shown here were obtained using this formulation with a GP prior on the growth rate.
@@
 We assumed that the probability of observing prevalence $Y_{\text{P}, t}$ at time $t$ was given by independent normal distributions with mean $P(t)$ and standard deviation
-$$\sigma_{\text{P}, t} = \sigma_\text{P} + Y^\sigma_{\text{P}, t}$$
-where $\sigma_\text{P}$ was estimated as part of the inference procedure and $Y^\sigma_{\text{P}, t}$ calculated based on the reported confidence intervals in the ONS data, assuming independent normal errors.
+$$\sigma_{\text{P}, t} = \sqrt{\sigma_\text{P}^2 + Y^\sigma_{\text{P}^2, t}}$$
+where $\sigma_\text{P}$ was estimated as part of the inference procedure and $Y^\sigma_{\text{P}, t}$ calculated based on the reported credible intervals in the ONS data, assuming independent normal errors.
 For data sets where only weekly estimates were reported by ONS, for example at the sub-regional level, we calculated average prevalence across the time period reported from our daily prevalence estimates.
@@
 
-Here, the additional parameter $\beta$ can be interpreted as proportion of new infections that does not increase the population proportion with antibodies, either due to lack of seroconversion or because they are breakthrough infections in those with existing antibodies, and parameters $k$ and $l$ govern the degree to which new seropostives preferentially arise in those not seropositive so far.
+Here, the additional parameter $\beta$ can be interpreted as proportion of new infections that does not increase the population proportion with antibodies, either due to lack of seroconversion or because they are breakthrough infections in those with existing antibodies, and parameters $k$ and $l$ govern the degree to which new seropositives preferentially arise in those not seropositive so far.
 Additional parameters $\gamma_\text{I}$ and $\gamma_\text{V}$ can be interpreted as rates of waning from natural infection and vaccination, respectively.

### 2022-03-06 docs/paper.Rmd
results

@@
   kbl(caption = paste0(
-    "Estimates and credible intervals (CIs, as quantiles of the posterior",
+    "Estimates and credible intervals (CIs, as quantiles of the posterior ",
     "distribution) of biological parameters."

### 2022-03-07 scripts/analyse.R
Paper (#32)
* update pre-commit

* dep update to get report rendering

* check paper renders

* read data read in

* use data.table, build pkg docs, switch to using load_all() everywhere

* abstract

* introduction

* methods

* results

* discussion

* README

@@ -38,8 +38,7 @@ age <- !is.null(opts$age) && opts$age
 variants <- !is.null(opts$variants) && opts$variants
 
 ## Get tools
-functions <- list.files(here("R"), full.names = TRUE)
-walk(functions, source)
+devtools::load_all()
 
 if (local) {
   suffix <- "_local"

### 2022-03-07 scripts/plot_data.r

@@ -5,8 +5,7 @@ library(socialmixr)
 library(dplyr)
 
 ## Get tools
-functions <- list.files(here("R"), full.names = TRUE)
-walk(functions, source)
+devtools::load_all()
 
 prev <- read_cis()
 

### 2022-03-07 docs/paper.Rmd

@@
 
-We have presented a method to estimate epidemiological parameters such as infection incidence, time-varying reproduction numbers and growth rates from repeated cross-sectional PCR positivity estimates reported.
+We have presented a method to estimate epidemiological parameters such as infection incidence, time-varying reproduction numbers and growth rates from repeated cross-sectional PCR positivity estimates.
 The estimates of infection incidence are distinct from estimates of PCR positivity incidence that are reported alongside the positivity prevalence estimates, as the probability of detecting infections is low early in the course of an infection, and more generally varies over said course [@hellewell2021].
-When additionally using antibody and vaccination data, we recover estimates of relevant parameters such as seroconversion and waning rates that can be used to estimate antibody prevalence where infection and vaccination data is available but antibody data is not.
+When additionally using antibody and vaccination data, we refine our estimates of infection incidence and recover estimates of relevant parameters such as seroconversion and waning rates that can be used to estimate antibody prevalence where infection and vaccination data is available but antibody data is not.
 
@@
 Future directions of this work should help address some of these limitations, for example by including more detail on antibody levels, or by including antibody measurements that may be able to distinguish between natural and vaccine-acquired immunity [@amjadi2021].
-It could further make use of more comprehensive information on PCR detection curves taking into account correlations in detectability since time from infection.
+It could further make use of more comprehensive information on PCR detection curves taking into account correlations in detectability since time from infection and pointly jointly estimating these curves using individual level data.
 Combined with other data streams, for example on test-positive community cases, or severe outcomes resulting in hospitalisations or deaths, our method could be used to understand rates of notification or sever disease given infection, or to generate forecasts of expected burden.

### 2022-03-09 R/postprocess.R
postprocess inc vars

@@ -62,8 +62,11 @@ i2p_add_var <- function(dt, prev, data, var_col = NULL) {
       ,
       paste(var_col) := fcase(
         name %in% c("est_prev", "est_ab", "infections",
-		    "dcases", "dab", "gen_dab", "r", "R"), 
-			      vars[n_index],
+		    "dcases", "dab", "gen_dab", "r", "R",
+		    "eta", "init_growth"),
+	vars[n_index],
+        name %in% c("rho", "alpha", "init_inc"),
+        vars[p_index],
         name %in% c("beta", "gamma", "delta", "k", "l"),
         paste0(vars[ab_index], collapse = ";")
       )

### 2022-03-09 R/simulate.R
generate local antibodies

@@ -1,0 +1,90 @@
+i2p_simulate <- function(dat, param, nsamples = NULL, type = "estimate") {
+
+  type <- match.arg(type, choices = c("estimate", "observations"))
+  files <- c("ab.stan", "gaussian_process.stan",
+             "observed_in_window.stan", "convolve.stan")
+
+  functions <-
+    paste0("\n functions{ \n", paste(purrr::map_chr(
+      files, ~paste(readLines(here::here("stan", "functions", .)),
+                    collapse = "\n")),
+      collapse = "\n"), "\n }")
+
+  ## fix rstan / cmdstanr compatibility issue
+  functions <- gsub("array\\[\\] ([a-z]+) ([a-z_]+)", "\\1[] \\2", functions)
+
+  rstan::expose_stan_functions(rstan::stanc(model_code = functions))
+
+  ## transformed data
+
+  PHI <- setup_gp(dat$M, dat$L, dat$t - dat$diff_order)
+  vacc_with_ab <- t(apply(dat$vacc, 1, convolve, dat$vacc_ab_delay))
+
+  gp <- lapply(seq_len(dat$n), function(i) {
+    upd <- update_gp(PHI, dat$M, dat$L, param$alpha[i], param$rho[i], 
+		     param$eta[i, ], 0)
+    return(to_row_vector(upd))
+  })
+  gp <- do.call(rbind, gp)
+
+  infections <- lapply(seq_len(dat$n), function(i) {
+    inf <- plogis(param$init_inc[i] + gp[i, ])
+    return(to_row_vector(inf))
+  })
+  infections <- do.call(rbind, infections)
+
+  dcases <- lapply(seq_len(dat$n), function(i) {
+    dca <- convolve(infections[i, ], param$prob_detect)
+    return(to_row_vector(dca))
+  })
+  dcases <- do.call(rbind, dcases)
+
+  odcases <- lapply(seq_len(dat$n), function(i) {
+    odc <- observed_in_window(dcases[i, ], dat$prev_stime, dat$prev_etime, 
+			      dat$ut, dat$obs)
+    return(to_row_vector(odc))
+  })
+  odcases <- do.call(rbind, odcases)
+
+  combined_sigma <- lapply(seq_len(dat$n), function(i) {
+    cs <- sqrt(rep(param$sigma^2, ncol(dat$prev_sd2)) + dat$prev_sd2[i, ])
+    return(to_row_vector(cs))
+  })
[…]

### 2022-03-09 scripts/generate_local_ab.r

@@ -1,0 +1,34 @@
+library("here")
+library("data.table")
+library("tidyr")
+
+samples <- readRDS(here::here("outputs", "samples_regional_ab.rds"))
+diag <- readRDS(here::here("outputs", "diagnostics_regional_ab.rds"))
+dat <- diag$data[[1]]
+
+## create param and fix if desired
+get_params <- function(dt, fix = list()) {
+  dt <- dt[[1]]
+  ret <- list()
+  for (loop_name in unique(dt$name)) {
+    name_dt <- dt[name == loop_name]
+    indices <- 
+      vapply(grep("_index$", colnames(name_dt), value = TRUE),
+             function(type) {
+	       ifelse(any(is.na(name_dt[[type]])), 0, max(name_dt[[type]]))
+	     }, 0)
+    present_indices <- names(indices)[indices > 0]
+    ret[[loop_name]] <- array(name_dt$value, dim = indices[present_indices])
+  }
+
+  for (loop_name in names(fix)) {
+    ret[[loop_name]][] <- fix[[loop_name]]
+  }
+    
+  return(list(ret))
+}
+
+samples <- nest(samples, data = !sample)
+samples <- samples[, list(param = get_params(data)), by = sample]
+sims <- samples[, list(sim = i2p_simulate(dat, param[[1]])), by = sample]
+

### 2022-03-17 docs/paper.Rmd
minor paper edits; waning estimates now monthly (#33)

@@
 csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
-date: "`r format(Sys.Date(), format = '%B %d, %Y')` - WORK IN PROGRESS DRAFT"
+date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
 output:
@@
 params <- fread(params_table)
+params <- melt(params, id.vars = c("name", "index"))
+params[name == "gamma", value := 1 - exp(-365.25 / 12 * value)]
+params <- dcast(params, name + index ~ variable)
 params[name == "gamma",
@@
     name == "beta", "Proportion infected that seroconvert",
-    name == "gamma (infection)", "Antibody waning following infection (daily rate)",
-    name == "gamma (vaccination)", "Antibody waning following vaccination (daily rate)",
+    name ==
+    "gamma (infection)", "Antibody waning following infection (per month)",
+    name ==
+    "gamma (vaccination)", "Antibody waning following vaccination (per month)",
     name == "delta", "Proportion vaccinated that seroconvert",
@@
 Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022].
-We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, compared to 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020].
+We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, consistent with the 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020].
 We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
-Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (infection)"]$q95, 2)`--`r signif(1 / params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(1 / params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.
+Our estimates of waning suggest that detectable antibodies decrease by `r signif(params[name == "gamma (infection)"]$median, 2) * 100`% (90% CI: `r signif(params[name == "gamma (infection)"]$q5, 2) * 100`--`r signif(params[name == "gamma (infection)"]$q95, 2) * 100`) following infection and `r signif(params[name == "gamma (vaccination)"]$median, 2) * 100`% (90% CI: `r signif(params[name == "gamma (vaccination)"]$q5, 2) * 100`--`r signif(params[name == "gamma (vaccination)"]$q95, 2) * 100`) following vaccination.
+All these values depend on the specific cutoff used for seropositivity and combine a range of vaccines, and they ignore additional effects from receiving multiple doses of vaccine, becoming infected as well as vaccinated.
+They cannot be compared directly to estimates of vaccine efficacy or waning thereof.
 

### 2022-03-28 R/model.R
better inits for ab model

@@ -263,9 +263,9 @@ i2p_inits <- function(dat) {
 
     if (dat$ab_obs > 0) {
       init_list <- c(init_list, list(
-        beta = array(inv_logit(rnorm(1, -2, 0.4))),
-        gamma = array(inv_logit(rnorm(2, -9, 0.4))),
-        delta = array(inv_logit(rnorm(1, 3, 0.4))),
+        beta = array(inv_logit(rnorm(1, 2, 1))),
+        gamma = array(inv_logit(rnorm(2, -9, 4))),
+        delta = array(inv_logit(rnorm(1, 3, 1))),
         k = array(exp(rnorm(1, 0, 0.1))),
         l = array(exp(rnorm(1, 0, 0.1))),
         ab_sigma = array(

### 2022-03-29 docs/paper.Rmd
correct typo

@@
 
-We used a guassian process to model incidence of infections and then estimated observed PCR prevalence by convolving our modelled incidence estimates with a previously published PCR detection curve describing the probability of a positive test as a function of the time since infection.
+We used a gaussian process to model incidence of infections and then estimated observed PCR prevalence by convolving our modelled incidence estimates with a previously published PCR detection curve describing the probability of a positive test as a function of the time since infection.
 We refined our incidence estimates using time-varying estimates of antibody prevalence combined with a model of antibody positivity and waning that moved individuals between compartments with or without antibodies based on estimates of new infections, vaccination, probability of seroconversion and waning.
